---
title: "Data"
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
library(readr)
library(dplyr)
library(reactable)

rtheme <- reactableTheme(
  borderColor    = "rgba(23,21,15,0.12)",
  highlightColor = "#f0ece2",
  headerStyle    = list(borderBottom = "1.5px solid #17150f", fontWeight = 600),
  style          = list(fontFamily = '"IBM Plex Sans", system-ui, sans-serif',
                        fontSize = "13px"))
```

Every dataset here is public and free, with no login and no paid access. Where a
source would normally require accepting terms or running a manual download, this
project avoids it by using a small table stored with the project instead, and the
notes below say exactly what to replace if you want to extend the project.

## Base geography

Country shapes and coastlines come from **Natural Earth**, loaded through the
`rnaturalearth` package. Natural Earth is in the public domain, and because the
package ships the data, there is nothing to download by hand.

## Energy infrastructure

The infrastructure layer is a small table of the region's LNG import terminals,
`data/processed/lng_terminals.csv`. It holds the name, country, location, annual
import capacity (in billion cubic metres, or bcm), status, and a source note for
each terminal. You can sort or search it here:

```{r terminals-table}
#| echo: false
read_csv("data/processed/lng_terminals.csv", show_col_types = FALSE) |>
  transmute(Terminal = name, Country = country,
            `Capacity (bcm)` = capacity_bcm, Status = status,
            Type = type, Source = source) |>
  reactable(sortable = TRUE, searchable = TRUE, highlight = TRUE,
            borderless = TRUE, pagination = FALSE, theme = rtheme,
            columns = list(Source = colDef(minWidth = 220)))
```

The capacities are taken from operator reporting and the **Global Energy
Monitor** terminal tracker, current as of 2024 to 2025. The `source` column in
the table names the operator behind each figure. Two points worth flagging:

- Capacities move. Świnoujście reached 8.3 bcm only after its third tank came
  online in early 2025, and Mukran's 13.5 bcm assumes both of its FSRUs — the
  moored ships that store and re-gasify LNG — are running.
- The Paldiski jetty in Estonia is deliberately left out. It is shelved, and its
  capacity figure refers to the same ship (the Exemplar) that now operates at
  Inkoo, so including both would count one vessel twice.

To extend the project, **EMODnet Human Activities** and **OpenStreetMap** (via
the `osmdata` package) are the route to add the pipelines and cables as mapped
routes, and the full set of capacities can be reconciled against the **Global
Energy Monitor** tracker.

## Transit corridors

The corridor map on the Analysis page uses a small table,
`data/processed/transit_lines.csv`, listing the main subsea pipelines and cables
by their landing points and status. They are drawn as **schematic straight
lines**, not surveyed routes: the aim is to show which corridors exist and which
have been disrupted, not their exact path. Positions and dates come from operator
and press reporting. To map the true routes, add the actual route lines from
**EMODnet Human Activities** or **OpenStreetMap**.

## Proximity reference

Instead of a maritime boundary line, the proximity measure uses distance to the
nearest **Russian territory**, meaning the mainland near the Gulf of Finland plus
the Kaliningrad exclave, taken from Natural Earth. It is reproducible from a
public package and easy to read: how far each terminal is from Russian soil.

## Dependence

The dependence layer uses gas import dependency from **Eurostat** (indicator
`nrg_ind_id`, natural gas, 2023), held in `data/processed/import_dependence.csv`.
The figure is the share of gas demand met by net imports. Values above 100%
reflect stock draws or re-exports in that year, and Denmark is low because of its
own North Sea production. `scripts/03_eurostat.R` reproduces the pull.

```{r dependence-table}
#| echo: false
read_csv("data/processed/import_dependence.csv", show_col_types = FALSE) |>
  transmute(Country = country,
            `Import dependency (%)` = gas_import_dependency_pct) |>
  arrange(desc(`Import dependency (%)`)) |>
  reactable(sortable = TRUE, highlight = TRUE, borderless = TRUE,
            pagination = FALSE, theme = rtheme)
```

## A note on conflict data

This project frames the question as exposure and structural dependence, not
active conflict, because there is no battlefield on the Baltic seabed. ACLED is
deliberately not used, since it now requires registration, which would break the
public data rule. If a later version needs conflict events, UCDP GED is the
public, no-login substitute.

## How the data is prepared

The steps that turn the raw tables and map data into the layers behind the maps
live in `scripts/01_clean.R`, and the Analysis page repeats them when it builds:

- tidy up the columns,
- put everything on a map projection suited to Europe, so distances come out in
  kilometres rather than degrees,
- trim the map to the Baltic study area,
- keep the capacity and country information attached to each terminal.

## Citations

- Natural Earth. Free vector and raster map data, 1:50m cultural vectors.
  <https://www.naturalearthdata.com> (accessed June 2026).
- Eurostat. Energy import dependency by products, indicator `nrg_ind_id`,
  natural gas, 2023. <https://ec.europa.eu/eurostat> (accessed June 2026).
- Global Energy Monitor. Global LNG Terminal Tracker.
  <https://globalenergymonitor.org> (accessed June 2026).
- Operator figures for capacity and status: Gaz-System, Deutsche ReGas, Gasgrid
  Finland, KN Energies, and Hamina LNG (2024 to 2025 reporting).

If you add pipelines or cables later: EMODnet Human Activities
(<https://emodnet.ec.europa.eu/en/human-activities>) and OpenStreetMap.
