---
title: "Analysis"
---

This page computes everything from public data at render time. Country geometry
comes from Natural Earth (via the `rnaturalearth` package), and the terminal and
dependence tables are the small committed files in `data/processed/`. All
distance work is done in a projection suited to Europe (ETRS89 / LAEA Europe,
EPSG:3035), and the results are reported in kilometres.

The terminal capacities are rough approximations from public reporting, as
explained on the [Data](01-data.qmd) page. Read the rankings as a rough picture
of exposure, not authoritative figures.

```{r setup}
#| include: false
#| message: false
#| warning: false
library(sf)
library(dplyr)
library(readr)
library(ggplot2)
library(leaflet)
library(rnaturalearth)
library(scales)
library(knitr)

laea <- 3035

# country borders from Natural Earth
world <- ne_countries(scale = "medium", returnclass = "sf")

# study area: box around the Baltic
aoi <- st_as_sfc(st_bbox(c(xmin = 9, ymin = 53, xmax = 30, ymax = 66),
                         crs = 4326)) |>
  st_transform(laea)

# Russian territory, including Kaliningrad
russia <- world |>
  filter(admin == "Russia") |>
  st_transform(laea) |>
  st_make_valid()

# Danish Straits chokepoint
chokepoint <- st_as_sfc(st_bbox(c(xmin = 10.5, ymin = 55.0, xmax = 13.0, ymax = 56.2),
                                crs = 4326)) |>
  st_transform(laea)

terminals <- read_csv("data/processed/lng_terminals.csv", show_col_types = FALSE) |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  st_transform(laea)

# distances in km
terminals <- terminals |>
  mutate(
    dist_russia_km     = as.numeric(st_distance(geometry, st_union(russia)))     / 1000,
    dist_chokepoint_km = as.numeric(st_distance(geometry, st_union(chokepoint))) / 1000
  )

# exposure index weights
w_proximity  <- 0.4   # closer to Russia, higher
w_capacity   <- 0.3   # larger capacity, higher
w_chokepoint <- 0.3   # closer to the straits, higher

norm <- function(x) (max(x, na.rm = TRUE) - x) /
                    (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))

terminals <- terminals |>
  mutate(
    s_proximity  = norm(dist_russia_km),         # closer = higher
    s_chokepoint = norm(dist_chokepoint_km),
    s_capacity   = capacity_bcm / max(capacity_bcm, na.rm = TRUE),
    exposure     = w_proximity  * s_proximity +
                   w_capacity   * s_capacity +
                   w_chokepoint * s_chokepoint
  ) |>
  arrange(desc(exposure))

# basemap clipped to the study area
land_aoi <- world |> st_transform(laea) |> st_crop(aoi)
```

## Concentration

Where the import infrastructure sits, sized by regasification capacity. Russian
territory is shaded for reference.

```{r map-concentration}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 7
ggplot() +
  geom_sf(data = land_aoi, fill = "grey95", colour = "grey80", linewidth = 0.2) +
  geom_sf(data = st_crop(russia, aoi), fill = "grey75", colour = "grey60") +
  geom_sf(data = terminals, aes(size = capacity_bcm, colour = exposure)) +
  scale_size_continuous(range = c(2, 9), name = "Capacity (bcm/yr)") +
  scale_colour_distiller(palette = "YlOrRd", direction = 1, name = "Exposure") +
  labs(title = "Baltic LNG import terminals",
       subtitle = "Point size is capacity; shaded area is Russian territory") +
  theme_minimal() +
  theme(axis.text = element_blank(), panel.grid = element_blank())
```

## Proximity

Distance from each terminal to the nearest Russian territory, meaning the
mainland near the Gulf of Finland or the Kaliningrad exclave, with the share of
total capacity sitting within a given band.

```{r proximity-buffer}
#| echo: false
total_cap <- sum(terminals$capacity_bcm, na.rm = TRUE)
share_within <- function(km) {
  near <- terminals |> filter(dist_russia_km <= km)
  round(100 * sum(near$capacity_bcm, na.rm = TRUE) / total_cap)
}
```

`r share_within(200)`% of total regasification capacity sits within 200 km of
Russian territory, and `r share_within(150)`% within 150 km.

```{r dist-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 4
terminals |>
  st_drop_geometry() |>
  ggplot(aes(x = reorder(name, dist_russia_km), y = dist_russia_km)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(x = NULL, y = "km to nearest Russian territory",
       title = "Proximity to Russian territory") +
  theme_minimal()
```

## Dependence

Gas import dependency by country (Eurostat `nrg_ind_id`, natural gas, 2023),
joined to country polygons. The figure is the share of gas demand met by net
imports, so values above 100% reflect stock draws or re-exports.

```{r dependence-map}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 7
dep <- read_csv("data/processed/import_dependence.csv", show_col_types = FALSE)

dependence <- land_aoi |>
  left_join(dep, by = c("postal" = "iso_a2"))

ggplot() +
  geom_sf(data = dependence, aes(fill = gas_import_dependency_pct),
          colour = "white", linewidth = 0.2) +
  scale_fill_distiller(palette = "Reds", direction = 1, na.value = "grey90",
                       name = "Gas import\ndependency (%)") +
  labs(title = "Gas import dependency, 2023 (Eurostat)") +
  theme_minimal() +
  theme(axis.text = element_blank(), panel.grid = element_blank())
```

## Exposure ranking

```{r ranking}
#| echo: false
terminals |>
  st_drop_geometry() |>
  transmute(
    Terminal   = name,
    Country    = country,
    Status     = status,
    `Capacity (bcm)`     = capacity_bcm,
    `km to Russia`       = round(dist_russia_km),
    `km to Straits`      = round(dist_chokepoint_km),
    Exposure   = round(exposure, 2)
  ) |>
  kable()
```

## Interactive map

```{r leaflet}
#| echo: false
terminals_wgs <- st_transform(terminals, 4326)
pal <- colorNumeric("YlOrRd", domain = terminals_wgs$exposure)

leaflet(terminals_wgs) |>
  addProviderTiles("CartoDB.Positron") |>
  addCircleMarkers(
    radius      = ~rescale(capacity_bcm, to = c(5, 15)),
    color       = ~pal(exposure),
    fillOpacity = 0.8, stroke = FALSE,
    label       = ~paste0(name, ": exposure ", round(exposure, 2))
  ) |>
  addLegend("bottomright", pal = pal, values = ~exposure, title = "Exposure")
```
