---
title: "Analysis"
---

```{r setup}
#| include: false
#| eval: false
library(sf)
library(dplyr)
library(readr)
library(ggplot2)
library(tmap)
library(leaflet)
library(stringr)

# Processed layers are written by scripts/01_clean.R and committed to the repo,
# so this page renders without re-running the analysis or touching the network.
proc <- "data/processed"
```

## Area of interest

All measurements are in ETRS89 / LAEA Europe (EPSG:3035) so distances are in
metres, not degrees.

```{r aoi}
#| eval: false
aoi        <- st_read(file.path(proc, "aoi.gpkg"))
terminals  <- st_read(file.path(proc, "lng_terminals.gpkg"))
boundary   <- st_read(file.path(proc, "ru_maritime_boundary.gpkg"))
chokepoint <- st_read(file.path(proc, "danish_straits.gpkg"))
```

## Concentration

Where the infrastructure sits, sized by capacity.

```{r map-concentration}
#| eval: false
tm_shape(aoi) +
  tm_borders() +
  tm_shape(terminals) +
  tm_symbols(size = "capacity", col = "firebrick", alpha = 0.7) +
  tm_layout(title = "Baltic LNG terminals by capacity")
```

## Proximity

Distance from each terminal to the nearest Russian/Kaliningrad maritime
boundary, and to the Danish Straits gate. Buffers show how much capacity sits
within a given band of the boundary.

```{r proximity}
#| eval: false
terminals <- terminals |>
  mutate(
    dist_boundary_km   = as.numeric(st_distance(geom, st_union(boundary)))   / 1000,
    dist_chokepoint_km = as.numeric(st_distance(geom, st_union(chokepoint))) / 1000
  )

# share of total capacity within 50 km of the boundary
within_50 <- terminals |>
  filter(dist_boundary_km <= 50) |>
  summarise(cap = sum(capacity, na.rm = TRUE)) |>
  pull(cap)

total_cap <- sum(terminals$capacity, na.rm = TRUE)
share_50  <- within_50 / total_cap
```

```{r dist-distribution}
#| eval: false
ggplot(terminals, aes(x = reorder(name, dist_boundary_km), y = dist_boundary_km)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "km to Russian/Kaliningrad maritime boundary")
```

## Dependence

National energy import dependence, joined to country polygons.

```{r dependence}
#| eval: false
dependence <- st_read(file.path(proc, "dependence.gpkg"))

tm_shape(dependence) +
  tm_polygons(col = "import_dependency", palette = "Reds",
              title = "Import dependency (%)")
```

## Exposure index

Three normalized components, combined with stated weights. Keep the weights here
so a reader can see exactly what the index is doing.

```{r index}
#| eval: false
# weights — stated, defensible, easy to change
w_proximity   <- 0.4   # closer to the boundary -> higher exposure
w_capacity    <- 0.3   # larger share of national supply -> higher exposure
w_chokepoint  <- 0.3   # more reliant on the Danish Straits -> higher exposure

norm <- function(x) (max(x, na.rm = TRUE) - x) /
                    (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))

terminals <- terminals |>
  mutate(
    s_proximity  = norm(dist_boundary_km),      # inverted: nearer = higher
    s_chokepoint = 1 - norm(dist_chokepoint_km),
    s_capacity   = capacity / max(capacity, na.rm = TRUE),
    exposure     = w_proximity  * s_proximity +
                   w_capacity   * s_capacity +
                   w_chokepoint * s_chokepoint
  ) |>
  arrange(desc(exposure))

terminals |>
  st_drop_geometry() |>
  select(name, country, capacity, dist_boundary_km, exposure)
```

## Interactive map

```{r leaflet}
#| eval: false
terminals_wgs <- st_transform(terminals, 4326)

leaflet(terminals_wgs) |>
  addProviderTiles("CartoDB.Positron") |>
  addCircleMarkers(
    radius = ~scales::rescale(capacity, to = c(4, 14)),
    color  = ~ifelse(exposure > 0.6, "firebrick", "steelblue"),
    label  = ~paste0(name, " — exposure ", round(exposure, 2))
  )
```
