---
title: "Analysis"
---

This page computes everything from public data at render time. Country geometry
comes from Natural Earth (via the `rnaturalearth` package), and the terminal and
dependence tables are the small committed files in `data/processed/`. All
distance work is done in a projection suited to Europe (ETRS89 / LAEA Europe,
EPSG:3035), and the results are reported in kilometres.

The terminal capacities come from operator reporting and the Global Energy
Monitor tracker, current to 2024 and 2025 (see the [Data](01-data.qmd) page).
Capacities change, so read the rankings as a picture of exposure rather than
precise figures.

```{r setup}
#| include: false
#| message: false
#| warning: false
library(sf)
library(dplyr)
library(readr)
library(ggplot2)
library(leaflet)
library(rnaturalearth)
library(scales)
library(gt)
library(reactable)

laea <- 3035

# shared styling (editorial: warm paper, ink, oxblood accent)
ink   <- "#17150f"
muted <- "#6b6760"
exp_cols <- c("#e7e2d6", "#cf8a4e", "#9e2b25")  # low to high exposure, single warm ramp
theme_transit <- function(base = 12) {
  theme_minimal(base_size = base) +
    theme(
      plot.title    = element_text(face = "bold", colour = ink, size = 15),
      plot.subtitle = element_text(colour = muted, size = 11, margin = margin(b = 8)),
      plot.caption  = element_text(colour = muted, size = 8, hjust = 0,
                                   margin = margin(t = 10)),
      legend.title  = element_text(size = 9, colour = muted),
      legend.text   = element_text(size = 8),
      plot.background  = element_rect(fill = "#f7f4ec", colour = NA),
      panel.background = element_rect(fill = "#f7f4ec", colour = NA)
    )
}
theme_map <- function() {
  theme_transit() +
    theme(axis.text = element_blank(), axis.title = element_blank(),
          panel.grid = element_blank(),
          panel.background = element_rect(fill = "#eef0ec", colour = NA))
}

# country borders from Natural Earth
world <- ne_countries(scale = "medium", returnclass = "sf")

# study area: box around the Baltic
aoi <- st_as_sfc(st_bbox(c(xmin = 9, ymin = 53, xmax = 30, ymax = 66),
                         crs = 4326)) |>
  st_transform(laea)

# Russian territory, including Kaliningrad
russia <- world |>
  filter(admin == "Russia") |>
  st_transform(laea) |>
  st_make_valid()

# Danish Straits chokepoint
chokepoint <- st_as_sfc(st_bbox(c(xmin = 10.5, ymin = 55.0, xmax = 13.0, ymax = 56.2),
                                crs = 4326)) |>
  st_transform(laea)

terminals <- read_csv("data/processed/lng_terminals.csv", show_col_types = FALSE) |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  st_transform(laea)

# distances in km
terminals <- terminals |>
  mutate(
    dist_russia_km     = as.numeric(st_distance(geometry, st_union(russia)))     / 1000,
    dist_chokepoint_km = as.numeric(st_distance(geometry, st_union(chokepoint))) / 1000
  )

# exposure index weights
w_proximity  <- 0.4   # closer to Russia, higher
w_capacity   <- 0.3   # larger capacity, higher
w_chokepoint <- 0.3   # longer voyage past the straits, higher

norm <- function(x) (max(x, na.rm = TRUE) - x) /
                    (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))

terminals <- terminals |>
  mutate(
    s_proximity  = norm(dist_russia_km),             # closer = higher
    s_chokepoint = 1 - norm(dist_chokepoint_km),     # longer Baltic transit = higher
    s_capacity   = capacity_bcm / max(capacity_bcm, na.rm = TRUE),
    exposure     = w_proximity  * s_proximity +
                   w_capacity   * s_capacity +
                   w_chokepoint * s_chokepoint
  ) |>
  arrange(desc(exposure))

# distance band from Russian territory, for the proximity-zone map
terminals <- terminals |>
  mutate(zone = cut(dist_russia_km, breaks = c(0, 100, 200, 300, Inf),
                    labels = c("Under 100 km", "100 to 200 km",
                               "200 to 300 km", "Over 300 km")))

# basemap clipped to the study area
land_aoi <- world |> st_transform(laea) |> st_crop(aoi)

# transit corridors: schematic straight lines between landing points
tl <- read_csv("data/processed/transit_lines.csv", show_col_types = FALSE)
line_geom <- lapply(seq_len(nrow(tl)), function(i)
  st_linestring(matrix(c(tl$lon1[i], tl$lon2[i], tl$lat1[i], tl$lat2[i]), ncol = 2)))
transit <- st_sf(tl, geometry = st_sfc(line_geom, crs = 4326)) |>
  mutate(state = ifelse(status == "operating", "Operating", "Disrupted since 2022")) |>
  st_transform(laea)
corridor_cols <- c("Operating" = "#4a6a7a", "Disrupted since 2022" = "#9e2b25")
```

## Concentration

Where the import infrastructure sits, sized by regasification capacity. Russian
territory is shaded for reference.

```{r map-concentration}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 7
choke_sf <- st_sf(geometry = chokepoint)
ggplot() +
  geom_sf(data = land_aoi, fill = "#e7e2d6", colour = "#d6d0c2", linewidth = 0.2) +
  geom_sf(data = st_crop(russia, aoi), fill = "#e3c9c3", colour = "#cdaaa3",
          linewidth = 0.2) +
  geom_sf(data = choke_sf, fill = NA, colour = ink, linewidth = 0.5,
          linetype = "31") +
  geom_sf(data = terminals, aes(size = capacity_bcm, colour = exposure)) +
  geom_sf_text(data = terminals, aes(label = name), size = 2.6, colour = ink,
               nudge_y = 38000, check_overlap = TRUE) +
  geom_sf_text(data = choke_sf, aes(label = "Danish\nStraits"), size = 2.5,
               colour = ink) +
  scale_size_continuous(range = c(2.5, 11), name = "Capacity\n(bcm/yr)") +
  scale_colour_gradientn(colours = exp_cols, name = "Exposure") +
  labs(title = "Baltic LNG import terminals",
       subtitle = "Point size is capacity, colour is exposure; the dashed box is the Danish Straits",
       caption = "Gdansk is planned; the others are operating. Sources: Natural Earth, operators, and Global Energy Monitor.") +
  theme_map()
```

## Transit corridors

The terminals are only part of the picture. They replaced, or now sit alongside, a
network of subsea pipelines and cables, several of which have been damaged or
sabotaged since 2022. The map shows the main corridors as schematic straight
lines between their landing points.

```{r map-corridors}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 7
# label each corridor at its midpoint so labels do not collide
transit_lab <- suppressWarnings(st_centroid(transit))
ggplot() +
  geom_sf(data = land_aoi, fill = "#e7e2d6", colour = "#d6d0c2", linewidth = 0.2) +
  geom_sf(data = st_crop(russia, aoi), fill = "#e3c9c3", colour = "#cdaaa3",
          linewidth = 0.2) +
  geom_sf(data = terminals, colour = "#8a8578", size = 1.4) +
  geom_sf(data = transit, aes(colour = state), linewidth = 1) +
  geom_sf_text(data = transit_lab, aes(label = name), size = 2.5, colour = ink,
               nudge_y = 22000, check_overlap = TRUE) +
  scale_colour_manual(values = corridor_cols, name = NULL) +
  labs(title = "Major energy transit corridors",
       subtitle = "Grey points are the LNG terminals; red corridors have been damaged or sabotaged since 2022",
       caption = "Schematic straight lines between landing points, not surveyed routes. Sources: operator and press reporting.") +
  theme_map()
```

## Proximity

Distance from each terminal to the nearest Russian territory, meaning the
mainland near the Gulf of Finland or the Kaliningrad exclave, with the share of
total capacity sitting within a given band.

```{r proximity-buffer}
#| echo: false
op_cap <- sum(terminals$capacity_bcm[terminals$status == "operating"], na.rm = TRUE)
share_within <- function(km) {
  near <- terminals |> filter(status == "operating", dist_russia_km <= km)
  round(100 * sum(near$capacity_bcm, na.rm = TRUE) / op_cap)
}
```

Only `r share_within(200)`% of operating regasification capacity sits within 200 km
of Russian territory. The two largest terminals, Mukran and Świnoujście, lie more
than 350 km back, so the exposure concentrates in the smaller terminals closest to
the border.

```{r dist-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 4
terminals |>
  st_drop_geometry() |>
  ggplot(aes(x = reorder(name, -dist_russia_km), y = dist_russia_km)) +
  geom_col(fill = ink, width = 0.7) +
  geom_text(aes(label = round(dist_russia_km)), hjust = -0.2, size = 3,
            colour = muted) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "km to nearest Russian territory",
       title = "How close each terminal sits to Russian territory") +
  theme_transit() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank())
```

## Proximity zones

The rings mark distance from Russian territory in 100 km steps. The terminals
inside 100 km are the small ones, near Kaliningrad and at the eastern end of the
Gulf of Finland. The largest capacity sits in the outer band, well back from the
border.

```{r map-zones}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 7
ru_aoi <- st_make_valid(st_union(st_crop(russia, aoi)))
buf <- function(km) st_make_valid(st_intersection(st_buffer(ru_aoi, km * 1000), aoi))
b100 <- buf(100); b200 <- buf(200); b300 <- buf(300)
ggplot() +
  geom_sf(data = land_aoi, fill = "#e7e2d6", colour = "#d6d0c2", linewidth = 0.2) +
  geom_sf(data = b300, fill = "#9e2b25", alpha = 0.07, colour = "#c9a9a3", linewidth = 0.2) +
  geom_sf(data = b200, fill = "#9e2b25", alpha = 0.09, colour = "#c9a9a3", linewidth = 0.2) +
  geom_sf(data = b100, fill = "#9e2b25", alpha = 0.11, colour = "#c9a9a3", linewidth = 0.2) +
  geom_sf(data = st_crop(russia, aoi), fill = "#cc9a92", colour = "#b98980",
          linewidth = 0.2) +
  geom_sf(data = terminals, aes(size = capacity_bcm), colour = "#17150f") +
  geom_sf_text(data = terminals, aes(label = name), size = 2.5, colour = ink,
               nudge_y = 38000, check_overlap = TRUE) +
  scale_size_continuous(range = c(2, 9), name = "Capacity\n(bcm/yr)") +
  labs(title = "Distance from Russian territory",
       subtitle = "Rings at 100, 200 and 300 km; darker shading is closer to Russia",
       caption = "Buffers around Russian territory near the Baltic. Source: Natural Earth.") +
  theme_map()
```

```{r zone-table}
#| echo: false
terminals |>
  st_drop_geometry() |>
  group_by(Band = zone) |>
  summarise(Terminals = dplyr::n(),
            `Capacity (bcm)` = round(sum(capacity_bcm), 1), .groups = "drop") |>
  gt() |>
  tab_options(table.font.size = px(13), column_labels.font.weight = "600",
              column_labels.border.top.style = "none",
              column_labels.border.bottom.color = "#17150f",
              column_labels.border.bottom.width = px(1.5),
              table_body.hlines.color = "rgba(23,21,15,0.12)",
              table.border.top.style = "none", table.border.bottom.style = "none")
```

## Dependence

Gas import dependency by country (Eurostat `nrg_ind_id`, natural gas, 2023),
joined to country polygons. The figure is the share of gas demand met by net
imports, so values above 100% reflect stock draws or re-exports.

```{r dependence-map}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 7
dep <- read_csv("data/processed/import_dependence.csv", show_col_types = FALSE)

# Join on ISO-2, not Natural Earth's "postal" field: postal uses non-ISO codes
# for several countries (Germany "D", Sweden "S", Finland "FIN"), which silently
# dropped them from the map.
dependence <- land_aoi |>
  left_join(dplyr::select(dep, iso_a2, gas_import_dependency_pct),
            by = "iso_a2")

ggplot() +
  geom_sf(data = dependence, aes(fill = gas_import_dependency_pct),
          colour = "white", linewidth = 0.3) +
  scale_fill_gradient(low = "#efe7d8", high = "#9e2b25", na.value = "#e6e2d8",
                      name = "Import\ndependency (%)") +
  labs(title = "Gas import dependency by country, 2023",
       subtitle = "Share of gas demand met by net imports (Eurostat nrg_ind_id)",
       caption = "Uncoloured countries are outside the study set. Values above 100% reflect stock draws or re-exports.") +
  theme_map()
```

## Exposure ranking

Click any column heading to re-sort the table.

```{r ranking}
#| echo: false
rank_df <- terminals |>
  st_drop_geometry() |>
  transmute(
    Terminal   = name,
    Country    = country,
    Status     = status,
    Capacity   = round(capacity_bcm, 2),
    `km to Russia`  = round(dist_russia_km),
    `km to Straits` = round(dist_chokepoint_km),
    Exposure   = round(exposure, 2)
  )

exp_rng  <- range(rank_df$Exposure)
exp_ramp <- colorRamp(exp_cols)

reactable(
  rank_df,
  defaultSorted = list(Exposure = "desc"),
  sortable = TRUE, highlight = TRUE, borderless = TRUE, pagination = FALSE,
  columns = list(
    Capacity        = colDef(name = "Capacity (bcm)", align = "center"),
    `km to Russia`  = colDef(align = "center"),
    `km to Straits` = colDef(align = "center"),
    Exposure = colDef(align = "center", style = function(value) {
      x <- (value - exp_rng[1]) / (exp_rng[2] - exp_rng[1])
      list(background  = grDevices::rgb(exp_ramp(x), maxColorValue = 255),
           color       = if (x > 0.55) "#ffffff" else "#17150f",
           fontWeight  = 600)
    })
  ),
  theme = reactableTheme(
    borderColor    = "rgba(23,21,15,0.12)",
    highlightColor = "#f0ece2",
    headerStyle    = list(borderBottom = "1.5px solid #17150f", fontWeight = 600),
    style          = list(fontFamily = '"IBM Plex Sans", system-ui, sans-serif',
                          fontSize = "13px")
  )
)
```

## How sensitive is the ranking?

The weights are a choice, so it helps to see how much they change the result. The
table recomputes the score under three weightings: balanced (the version used
above), proximity-led, and capacity-led.

```{r sensitivity}
#| echo: false
scen <- function(wp, wc, wk) {
  d <- st_drop_geometry(terminals)
  wp * d$s_proximity + wc * d$s_capacity + wk * d$s_chokepoint
}
st_drop_geometry(terminals) |>
  transmute(Terminal = name,
            Balanced        = round(scen(0.4, 0.3, 0.3), 2),
            `Proximity-led` = round(scen(0.6, 0.2, 0.2), 2),
            `Capacity-led`  = round(scen(0.2, 0.6, 0.2), 2)) |>
  arrange(desc(Balanced)) |>
  gt() |>
  tab_header(title = "Exposure score under three weightings") |>
  data_color(columns = c(Balanced, `Proximity-led`, `Capacity-led`),
             palette = exp_cols) |>
  cols_align(align = "center",
             columns = c(Balanced, `Proximity-led`, `Capacity-led`)) |>
  tab_options(table.font.size = px(13), heading.title.font.size = px(15),
              heading.align = "left", column_labels.font.weight = "600",
              column_labels.border.top.style = "none",
              column_labels.border.bottom.color = "#17150f",
              column_labels.border.bottom.width = px(1.5),
              table_body.hlines.color = "rgba(23,21,15,0.12)",
              table.border.top.style = "none",
              table.border.bottom.style = "none")
```

The terminals nearest Russian territory, Hamina, Klaipėda, and Gdańsk, lead under
the proximity-led weighting, while Mukran, much the largest, rises to the top
under the capacity-led one. The order depends on the weights, which are easy to
change.

## Interactive map

Toggle the layers in the top-right, hover a corridor or terminal for its name,
and click a terminal for its figures.

```{r leaflet}
#| echo: false
terminals_wgs <- st_transform(terminals, 4326)
russia_wgs    <- st_transform(st_crop(russia, aoi), 4326)
transit_wgs   <- st_transform(transit, 4326)
choke_bb      <- st_bbox(st_transform(chokepoint, 4326))
pal <- colorNumeric(exp_cols, domain = terminals_wgs$exposure)
tb  <- st_bbox(terminals_wgs)

leaflet(terminals_wgs) |>
  addProviderTiles("CartoDB.Positron") |>
  # Focus the initial view on the terminals. Without this, leaflet auto-fits to
  # all layers, and the Russia polygon stretches the view east into empty land.
  fitBounds(tb[["xmin"]] - 1.5, tb[["ymin"]] - 0.8,
            tb[["xmax"]] + 1.5, tb[["ymax"]] + 0.8) |>
  addPolygons(data = russia_wgs, fillColor = "#9e2b25", fillOpacity = 0.16,
              weight = 0, group = "Russian territory", label = "Russian territory",
              highlightOptions = highlightOptions(fillOpacity = 0.32)) |>
  addRectangles(lng1 = choke_bb[["xmin"]], lat1 = choke_bb[["ymin"]],
                lng2 = choke_bb[["xmax"]], lat2 = choke_bb[["ymax"]],
                fill = FALSE, color = "#17150f", weight = 2, dashArray = "5,5",
                group = "Danish Straits", label = "Danish Straits") |>
  addPolylines(data = transit_wgs, color = ~corridor_cols[state], weight = 3,
               opacity = 0.9, label = ~paste0(name, " (", status, ")"),
               highlightOptions = highlightOptions(weight = 6, opacity = 1),
               group = "Transit corridors") |>
  addCircleMarkers(
    radius      = ~rescale(capacity_bcm, to = c(5, 16)),
    color       = "#17150f", weight = 1,
    fillColor   = ~pal(exposure), fillOpacity = 0.85,
    group       = "LNG terminals",
    label       = ~name,
    popup       = ~paste0("<b>", name, "</b><br>", country,
                          "<br>Capacity: ", capacity_bcm, " bcm/yr",
                          "<br>Status: ", status,
                          "<br>Exposure: ", round(exposure, 2))
  ) |>
  addLegend("bottomright", pal = pal, values = ~exposure, title = "Exposure",
            opacity = 0.9) |>
  addScaleBar(position = "bottomleft") |>
  addMiniMap(toggleDisplay = TRUE, minimized = FALSE, width = 110, height = 110) |>
  addLayersControl(
    overlayGroups = c("LNG terminals", "Transit corridors",
                      "Russian territory", "Danish Straits"),
    options = layersControlOptions(collapsed = FALSE))
```
