---
title: "Analysis"
---

This page computes everything from public data at render time. Country geometry
comes from Natural Earth (via the `rnaturalearth` package), and the terminal and
dependence tables are the small committed files in `data/processed/`. All
distance work is done in a projection suited to Europe (ETRS89 / LAEA Europe,
EPSG:3035), and the results are reported in kilometres.

The terminal capacities come from operator reporting and the Global Energy
Monitor tracker, current to 2024 and 2025 (see the [Data](01-data.qmd) page).
Capacities change, so read the rankings as a picture of exposure rather than
precise figures.

```{r setup}
#| include: false
#| message: false
#| warning: false
library(sf)
library(dplyr)
library(readr)
library(ggplot2)
library(leaflet)
library(rnaturalearth)
library(scales)
library(gt)

laea <- 3035

# shared styling (editorial: warm paper, ink, oxblood accent)
ink   <- "#17150f"
muted <- "#6b6760"
exp_cols <- c("#e7e2d6", "#cf8a4e", "#9e2b25")  # low to high exposure, single warm ramp
theme_transit <- function(base = 12) {
  theme_minimal(base_size = base) +
    theme(
      plot.title    = element_text(face = "bold", colour = ink, size = 15),
      plot.subtitle = element_text(colour = muted, size = 11, margin = margin(b = 8)),
      plot.caption  = element_text(colour = muted, size = 8, hjust = 0,
                                   margin = margin(t = 10)),
      legend.title  = element_text(size = 9, colour = muted),
      legend.text   = element_text(size = 8),
      plot.background  = element_rect(fill = "#f7f4ec", colour = NA),
      panel.background = element_rect(fill = "#f7f4ec", colour = NA)
    )
}
theme_map <- function() {
  theme_transit() +
    theme(axis.text = element_blank(), axis.title = element_blank(),
          panel.grid = element_blank(),
          panel.background = element_rect(fill = "#eef0ec", colour = NA))
}

# country borders from Natural Earth
world <- ne_countries(scale = "medium", returnclass = "sf")

# study area: box around the Baltic
aoi <- st_as_sfc(st_bbox(c(xmin = 9, ymin = 53, xmax = 30, ymax = 66),
                         crs = 4326)) |>
  st_transform(laea)

# Russian territory, including Kaliningrad
russia <- world |>
  filter(admin == "Russia") |>
  st_transform(laea) |>
  st_make_valid()

# Danish Straits chokepoint
chokepoint <- st_as_sfc(st_bbox(c(xmin = 10.5, ymin = 55.0, xmax = 13.0, ymax = 56.2),
                                crs = 4326)) |>
  st_transform(laea)

terminals <- read_csv("data/processed/lng_terminals.csv", show_col_types = FALSE) |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  st_transform(laea)

# distances in km
terminals <- terminals |>
  mutate(
    dist_russia_km     = as.numeric(st_distance(geometry, st_union(russia)))     / 1000,
    dist_chokepoint_km = as.numeric(st_distance(geometry, st_union(chokepoint))) / 1000
  )

# exposure index weights
w_proximity  <- 0.4   # closer to Russia, higher
w_capacity   <- 0.3   # larger capacity, higher
w_chokepoint <- 0.3   # longer voyage past the straits, higher

norm <- function(x) (max(x, na.rm = TRUE) - x) /
                    (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))

terminals <- terminals |>
  mutate(
    s_proximity  = norm(dist_russia_km),             # closer = higher
    s_chokepoint = 1 - norm(dist_chokepoint_km),     # longer Baltic transit = higher
    s_capacity   = capacity_bcm / max(capacity_bcm, na.rm = TRUE),
    exposure     = w_proximity  * s_proximity +
                   w_capacity   * s_capacity +
                   w_chokepoint * s_chokepoint
  ) |>
  arrange(desc(exposure))

# basemap clipped to the study area
land_aoi <- world |> st_transform(laea) |> st_crop(aoi)
```

## Concentration

Where the import infrastructure sits, sized by regasification capacity. Russian
territory is shaded for reference.

```{r map-concentration}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 7
choke_sf <- st_sf(geometry = chokepoint)
ggplot() +
  geom_sf(data = land_aoi, fill = "#e7e2d6", colour = "#d6d0c2", linewidth = 0.2) +
  geom_sf(data = st_crop(russia, aoi), fill = "#e3c9c3", colour = "#cdaaa3",
          linewidth = 0.2) +
  geom_sf(data = choke_sf, fill = NA, colour = ink, linewidth = 0.5,
          linetype = "31") +
  geom_sf(data = terminals, aes(size = capacity_bcm, colour = exposure)) +
  geom_sf_text(data = terminals, aes(label = name), size = 2.6, colour = ink,
               nudge_y = 38000, check_overlap = TRUE) +
  geom_sf_text(data = choke_sf, aes(label = "Danish\nStraits"), size = 2.5,
               colour = ink) +
  scale_size_continuous(range = c(2.5, 11), name = "Capacity\n(bcm/yr)") +
  scale_colour_gradientn(colours = exp_cols, name = "Exposure") +
  labs(title = "Baltic LNG import terminals",
       subtitle = "Point size is capacity, colour is exposure; the dashed box is the Danish Straits",
       caption = "Gdansk is planned; the others are operating. Sources: Natural Earth, operators, and Global Energy Monitor.") +
  theme_map()
```

## Proximity

Distance from each terminal to the nearest Russian territory, meaning the
mainland near the Gulf of Finland or the Kaliningrad exclave, with the share of
total capacity sitting within a given band.

```{r proximity-buffer}
#| echo: false
op_cap <- sum(terminals$capacity_bcm[terminals$status == "operating"], na.rm = TRUE)
share_within <- function(km) {
  near <- terminals |> filter(status == "operating", dist_russia_km <= km)
  round(100 * sum(near$capacity_bcm, na.rm = TRUE) / op_cap)
}
```

`r share_within(200)`% of operating regasification capacity sits within 200 km of
Russian territory, and `r share_within(150)`% within 150 km.

```{r dist-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 4
terminals |>
  st_drop_geometry() |>
  ggplot(aes(x = reorder(name, -dist_russia_km), y = dist_russia_km)) +
  geom_col(fill = ink, width = 0.7) +
  geom_text(aes(label = round(dist_russia_km)), hjust = -0.2, size = 3,
            colour = muted) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "km to nearest Russian territory",
       title = "How close each terminal sits to Russian territory") +
  theme_transit() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank())
```

## Dependence

Gas import dependency by country (Eurostat `nrg_ind_id`, natural gas, 2023),
joined to country polygons. The figure is the share of gas demand met by net
imports, so values above 100% reflect stock draws or re-exports.

```{r dependence-map}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 7
dep <- read_csv("data/processed/import_dependence.csv", show_col_types = FALSE)

dependence <- land_aoi |>
  left_join(dplyr::select(dep, iso_a2, gas_import_dependency_pct),
            by = c("postal" = "iso_a2"))

ggplot() +
  geom_sf(data = dependence, aes(fill = gas_import_dependency_pct),
          colour = "white", linewidth = 0.3) +
  scale_fill_gradient(low = "#efe7d8", high = "#9e2b25", na.value = "#e6e2d8",
                      name = "Import\ndependency (%)") +
  labs(title = "Gas import dependency by country, 2023",
       subtitle = "Share of gas demand met by net imports (Eurostat nrg_ind_id)",
       caption = "Uncoloured countries are outside the study set. Values above 100% reflect stock draws or re-exports.") +
  theme_map()
```

## Exposure ranking

```{r ranking}
#| echo: false
terminals |>
  st_drop_geometry() |>
  transmute(
    Terminal   = name,
    Country    = country,
    Status     = status,
    Capacity   = capacity_bcm,
    `km to Russia`  = round(dist_russia_km),
    `km to Straits` = round(dist_chokepoint_km),
    Exposure   = round(exposure, 2)
  ) |>
  gt() |>
  tab_header(title = "Terminals ranked by exposure") |>
  fmt_number(columns = Capacity, decimals = 2) |>
  data_color(columns = Exposure, palette = exp_cols,
             domain = range(round(terminals$exposure, 2))) |>
  cols_label(Capacity = "Capacity (bcm)") |>
  cols_align(align = "center",
             columns = c(Capacity, `km to Russia`, `km to Straits`, Exposure)) |>
  tab_options(table.font.size = px(13), heading.title.font.size = px(15),
              heading.align = "left",
              column_labels.font.weight = "600",
              column_labels.border.top.style = "none",
              column_labels.border.bottom.color = "#17150f",
              column_labels.border.bottom.width = px(1.5),
              table_body.hlines.color = "rgba(23,21,15,0.12)",
              table.border.top.style = "none",
              table.border.bottom.style = "none")
```

## How sensitive is the ranking?

The weights are a choice, so it is worth seeing how much they move the result.
The table recomputes the score under three weightings: balanced (the version used
above), proximity-led, and capacity-led.

```{r sensitivity}
#| echo: false
scen <- function(wp, wc, wk) {
  d <- st_drop_geometry(terminals)
  wp * d$s_proximity + wc * d$s_capacity + wk * d$s_chokepoint
}
st_drop_geometry(terminals) |>
  transmute(Terminal = name,
            Balanced        = round(scen(0.4, 0.3, 0.3), 2),
            `Proximity-led` = round(scen(0.6, 0.2, 0.2), 2),
            `Capacity-led`  = round(scen(0.2, 0.6, 0.2), 2)) |>
  arrange(desc(Balanced)) |>
  gt() |>
  tab_header(title = "Exposure score under three weightings") |>
  data_color(columns = c(Balanced, `Proximity-led`, `Capacity-led`),
             palette = exp_cols) |>
  cols_align(align = "center",
             columns = c(Balanced, `Proximity-led`, `Capacity-led`)) |>
  tab_options(table.font.size = px(13), heading.title.font.size = px(15),
              heading.align = "left", column_labels.font.weight = "600",
              column_labels.border.top.style = "none",
              column_labels.border.bottom.color = "#17150f",
              column_labels.border.bottom.width = px(1.5),
              table_body.hlines.color = "rgba(23,21,15,0.12)",
              table.border.top.style = "none",
              table.border.bottom.style = "none")
```

The Gulf of Finland terminals climb under the proximity-led weighting, while
Mukran rises to the top under the capacity-led one. The order is a function of
the assumptions, and those assumptions are visible and easy to change.

## Interactive map

```{r leaflet}
#| echo: false
terminals_wgs <- st_transform(terminals, 4326)
russia_wgs    <- st_transform(st_crop(russia, aoi), 4326)
choke_bb      <- st_bbox(st_transform(chokepoint, 4326))
pal <- colorNumeric(exp_cols, domain = terminals_wgs$exposure)

leaflet(terminals_wgs) |>
  addProviderTiles("CartoDB.Positron") |>
  addPolygons(data = russia_wgs, fillColor = "#9e2b25", fillOpacity = 0.16,
              weight = 0, group = "Russian territory") |>
  addRectangles(lng1 = choke_bb[["xmin"]], lat1 = choke_bb[["ymin"]],
                lng2 = choke_bb[["xmax"]], lat2 = choke_bb[["ymax"]],
                fill = FALSE, color = "#17150f", weight = 2, dashArray = "5,5",
                group = "Danish Straits") |>
  addCircleMarkers(
    radius      = ~rescale(capacity_bcm, to = c(5, 16)),
    color       = "#17150f", weight = 1,
    fillColor   = ~pal(exposure), fillOpacity = 0.85,
    popup       = ~paste0("<b>", name, "</b><br>", country,
                          "<br>Capacity: ", capacity_bcm, " bcm/yr",
                          "<br>Status: ", status,
                          "<br>Exposure: ", round(exposure, 2))
  ) |>
  addLegend("bottomright", pal = pal, values = ~exposure, title = "Exposure",
            opacity = 0.9) |>
  addLayersControl(overlayGroups = c("Russian territory", "Danish Straits"),
                   options = layersControlOptions(collapsed = FALSE))
```
