---
title: "Analysis"
execute:
  warning: false
  message: false
---

Every map and figure on this page is built from the public data listed on the
[Data](01-data.qmd) page, and all distances are reported in kilometres. Each chart
has a "Show code" toggle if you want to see how it was made, but you do not need
to read any code to follow the analysis.

The terminal capacities come from operator reporting and the Global Energy
Monitor tracker, current to 2024 and 2025 (see the [Data](01-data.qmd) page).
Capacities change, so read the rankings as a picture of exposure rather than
precise figures.

```{r setup}
#| include: false
#| message: false
#| warning: false
library(sf)
library(dplyr)
library(readr)
library(ggplot2)
library(leaflet)
library(rnaturalearth)
library(scales)
library(gt)
library(reactable)
library(ggrepel)

laea <- 3035

# shared styling: cool neutrals, navy structure, red reserved for risk
ink   <- "#12263f"
muted <- "#5a6675"
land  <- "#e6ebf0"
land_edge <- "#d3dae1"
ru_fill   <- "#dfd0cf"
ru_edge   <- "#c9b4b2"
sea   <- "#f7f9fb"
exp_cols <- c("#e9eef2", "#cf8f6a", "#8f1d17")  # low to high exposure
rtheme <- reactableTheme(
  borderColor    = "rgba(18,38,63,0.13)",
  highlightColor = "#eef2f6",
  headerStyle    = list(borderBottom = "1.5px solid #12263f", fontWeight = 600),
  style          = list(fontFamily = '"Source Sans 3", system-ui, sans-serif',
                        fontSize = "13px"))
theme_transit <- function(base = 12) {
  theme_minimal(base_size = base) +
    theme(
      plot.title    = element_text(face = "bold", colour = ink, size = 15),
      plot.subtitle = element_text(colour = muted, size = 11, margin = margin(b = 8)),
      plot.caption  = element_text(colour = muted, size = 8, hjust = 0,
                                   margin = margin(t = 10)),
      legend.title  = element_text(size = 9, colour = muted),
      legend.text   = element_text(size = 8),
      plot.background  = element_rect(fill = "#ffffff", colour = NA),
      panel.background = element_rect(fill = "#ffffff", colour = NA)
    )
}
theme_map <- function() {
  theme_transit() +
    theme(axis.text = element_blank(), axis.title = element_blank(),
          panel.grid = element_blank(),
          panel.background = element_rect(fill = sea, colour = NA))
}

# country borders from Natural Earth
world <- ne_countries(scale = "medium", returnclass = "sf")

# study area: box around the Baltic
aoi <- st_as_sfc(st_bbox(c(xmin = 9, ymin = 53, xmax = 30, ymax = 66),
                         crs = 4326)) |>
  st_transform(laea)

# Russian territory, including Kaliningrad
russia <- world |>
  filter(admin == "Russia") |>
  st_transform(laea) |>
  st_make_valid()

# Danish Straits chokepoint
chokepoint <- st_as_sfc(st_bbox(c(xmin = 10.5, ymin = 55.0, xmax = 13.0, ymax = 56.2),
                                crs = 4326)) |>
  st_transform(laea)

terminals <- read_csv("data/processed/lng_terminals.csv", show_col_types = FALSE) |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  st_transform(laea)

# distances in km
terminals <- terminals |>
  mutate(
    dist_russia_km     = as.numeric(st_distance(geometry, st_union(russia)))     / 1000,
    dist_chokepoint_km = as.numeric(st_distance(geometry, st_union(chokepoint))) / 1000
  )

# exposure index weights
w_proximity  <- 0.4   # closer to Russia, higher
w_capacity   <- 0.3   # larger capacity, higher
w_chokepoint <- 0.3   # longer voyage past the straits, higher

norm <- function(x) (max(x, na.rm = TRUE) - x) /
                    (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))

terminals <- terminals |>
  mutate(
    s_proximity  = norm(dist_russia_km),             # closer = higher
    s_chokepoint = 1 - norm(dist_chokepoint_km),     # longer Baltic transit = higher
    s_capacity   = capacity_bcm / max(capacity_bcm, na.rm = TRUE),
    exposure     = w_proximity  * s_proximity +
                   w_capacity   * s_capacity +
                   w_chokepoint * s_chokepoint
  ) |>
  arrange(desc(exposure))

# distance band from Russian territory, for the proximity-zone map
terminals <- terminals |>
  mutate(zone = cut(dist_russia_km, breaks = c(0, 100, 200, 300, Inf),
                    labels = c("Under 100 km", "100 to 200 km",
                               "200 to 300 km", "Over 300 km")))

# basemap clipped to the study area
land_aoi <- world |> st_transform(laea) |> st_crop(aoi)

# transit corridors: schematic straight lines between landing points
tl <- read_csv("data/processed/transit_lines.csv", show_col_types = FALSE)
line_geom <- lapply(seq_len(nrow(tl)), function(i)
  st_linestring(matrix(c(tl$lon1[i], tl$lon2[i], tl$lat1[i], tl$lat2[i]), ncol = 2)))
transit <- st_sf(tl, geometry = st_sfc(line_geom, crs = 4326)) |>
  mutate(state = ifelse(status == "operating", "Operating", "Disrupted since 2022")) |>
  st_transform(laea)
corridor_cols <- c("Operating" = "#2c5d7c", "Disrupted since 2022" = "#8f1d17")
```

## Concentration

Where the import terminals sit, with each point sized by how much gas it can bring
in. Russian territory is shaded for reference.

```{r map-concentration}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 7
#| fig-alt: "Map of the Baltic Sea showing LNG import terminals as points sized by capacity and coloured by exposure, with Russian territory shaded and the Danish Straits marked by a dashed box."
choke_sf <- st_sf(geometry = chokepoint)
ggplot() +
  geom_sf(data = land_aoi, fill = land, colour = land_edge, linewidth = 0.2) +
  geom_sf(data = st_crop(russia, aoi), fill = ru_fill, colour = ru_edge,
          linewidth = 0.2) +
  geom_sf(data = choke_sf, fill = NA, colour = ink, linewidth = 0.5,
          linetype = "31") +
  geom_sf(data = terminals, aes(size = capacity_bcm, colour = exposure)) +
  geom_text_repel(data = terminals, aes(label = name, geometry = geometry),
                  stat = "sf_coordinates", size = 2.6, colour = ink, seed = 1,
                  min.segment.length = 0, segment.colour = "#8e9aa6",
                  segment.size = 0.2, box.padding = 0.4, max.overlaps = Inf) +
  geom_sf_text(data = choke_sf, aes(label = "Danish\nStraits"), size = 2.5,
               colour = ink) +
  scale_size_continuous(range = c(2.5, 11), name = "Capacity\n(bcm/yr)") +
  scale_colour_gradientn(colours = exp_cols, name = "Exposure") +
  labs(title = "Baltic LNG import terminals",
       subtitle = "Point size is capacity, colour is exposure; the dashed box is the Danish Straits",
       caption = "Gdansk is planned; the others are operating. Sources: Natural Earth, operators, and Global Energy Monitor.") +
  theme_map()
```

## Transit corridors

The terminals are only part of the picture. They replaced, or now sit alongside, a
network of subsea pipelines and cables, several of which have been damaged or
sabotaged since 2022. The map shows the main corridors as schematic straight
lines between their landing points.

```{r map-corridors}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 7
#| fig-alt: "Map of the Baltic showing the main subsea pipelines and cables as lines, coloured blue where operating and red where damaged or sabotaged since 2022, with LNG terminals as grey points."
# label each corridor at its midpoint so labels do not collide
transit_lab <- suppressWarnings(st_centroid(transit))
ggplot() +
  geom_sf(data = land_aoi, fill = land, colour = land_edge, linewidth = 0.2) +
  geom_sf(data = st_crop(russia, aoi), fill = ru_fill, colour = ru_edge,
          linewidth = 0.2) +
  geom_sf(data = terminals, colour = "#8e9aa6", size = 1.4) +
  geom_sf(data = transit, aes(colour = state), linewidth = 1) +
  geom_sf_text(data = transit_lab, aes(label = name), size = 2.5, colour = ink,
               nudge_y = 22000, check_overlap = TRUE) +
  scale_colour_manual(values = corridor_cols, name = NULL) +
  labs(title = "Major energy transit corridors",
       subtitle = "Grey points are the LNG terminals; red corridors have been damaged or sabotaged since 2022",
       caption = "Schematic straight lines between landing points, not surveyed routes. Sources: operator and press reporting.") +
  theme_map()
```

## Proximity

Distance from each terminal to the nearest Russian territory, meaning the
mainland near the Gulf of Finland or the Kaliningrad exclave, with the share of
total capacity sitting within a given band.

```{r proximity-buffer}
#| echo: false
op_cap <- sum(terminals$capacity_bcm[terminals$status == "operating"], na.rm = TRUE)
share_within <- function(km) {
  near <- terminals |> filter(status == "operating", dist_russia_km <= km)
  round(100 * sum(near$capacity_bcm, na.rm = TRUE) / op_cap)
}
```

Only `r share_within(200)`% of the region's operating import capacity sits within
200 km of Russian territory. The two largest terminals, Świnoujście and Mukran,
lie more than 350 km back, so the exposure concentrates in the smaller terminals
closest to the border.

```{r dist-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 4
#| fig-alt: "Bar chart of each terminal's distance to the nearest Russian territory, ranging from Hamina at about 34 km to Mukran at about 388 km."
terminals |>
  st_drop_geometry() |>
  ggplot(aes(x = reorder(name, -dist_russia_km), y = dist_russia_km)) +
  geom_col(fill = ink, width = 0.7) +
  geom_text(aes(label = round(dist_russia_km)), hjust = -0.2, size = 3,
            colour = muted) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "km to nearest Russian territory",
       title = "How close each terminal sits to Russian territory") +
  theme_transit() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank())
```

## Proximity zones

The rings mark distance from Russian territory in 100 km steps. The terminals
inside 100 km are the small ones, near Kaliningrad and at the eastern end of the
Gulf of Finland. The largest capacity sits in the outer band, well back from the
border.

```{r map-zones}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 7
#| fig-alt: "Map of the Baltic with distance rings at 100, 200 and 300 km around Russian territory, shaded darker closer to Russia, showing which terminals fall in each band."
ru_aoi <- st_make_valid(st_union(st_crop(russia, aoi)))
buf <- function(km) st_make_valid(st_intersection(st_buffer(ru_aoi, km * 1000), aoi))
b100 <- buf(100); b200 <- buf(200); b300 <- buf(300)
ggplot() +
  geom_sf(data = land_aoi, fill = land, colour = land_edge, linewidth = 0.2) +
  geom_sf(data = b300, fill = "#8f1d17", alpha = 0.07, colour = "#c3b0ae", linewidth = 0.2) +
  geom_sf(data = b200, fill = "#8f1d17", alpha = 0.09, colour = "#c3b0ae", linewidth = 0.2) +
  geom_sf(data = b100, fill = "#8f1d17", alpha = 0.11, colour = "#c3b0ae", linewidth = 0.2) +
  geom_sf(data = st_crop(russia, aoi), fill = "#cbb0ad", colour = "#b39b98",
          linewidth = 0.2) +
  geom_sf(data = terminals, aes(size = capacity_bcm), colour = ink) +
  geom_text_repel(data = terminals, aes(label = name, geometry = geometry),
                  stat = "sf_coordinates", size = 2.5, colour = ink, seed = 1,
                  min.segment.length = 0, segment.colour = "#8e9aa6",
                  segment.size = 0.2, box.padding = 0.4, max.overlaps = Inf) +
  scale_size_continuous(range = c(2, 9), name = "Capacity\n(bcm/yr)") +
  labs(title = "Distance from Russian territory",
       subtitle = "Rings at 100, 200 and 300 km; darker shading is closer to Russia",
       caption = "Buffers around Russian territory near the Baltic. Source: Natural Earth.") +
  theme_map()
```

```{r zone-table}
#| echo: false
terminals |>
  st_drop_geometry() |>
  group_by(Band = zone) |>
  summarise(Terminals = dplyr::n(),
            `Capacity (bcm)` = round(sum(capacity_bcm), 1), .groups = "drop") |>
  gt() |>
  tab_options(table.font.size = px(13), column_labels.font.weight = "600",
              column_labels.border.top.style = "none",
              column_labels.border.bottom.color = "#12263f",
              column_labels.border.bottom.width = px(1.5),
              table_body.hlines.color = "rgba(18,38,63,0.13)",
              table.border.top.style = "none", table.border.bottom.style = "none")
```

## Dependence

Gas import dependency for each country (Eurostat `nrg_ind_id`, natural gas, 2023).
The figure is the share of a country's gas demand met by imports rather than its
own production, so values above 100% reflect stock draws or re-exports.

```{r dependence-map}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 7
#| fig-alt: "Choropleth map of Baltic countries shaded by gas import dependency in 2023, with most of the region near 100 percent and Denmark much lower."
dep <- read_csv("data/processed/import_dependence.csv", show_col_types = FALSE)

# Join on ISO-2, not Natural Earth's "postal" field: postal uses non-ISO codes
# for several countries (Germany "D", Sweden "S", Finland "FIN"), which silently
# dropped them from the map. Use full country shapes, not the clipped basemap, so
# Germany and Denmark are not cut off by the study box.
dep_map <- world |>
  st_transform(laea) |>
  left_join(dplyr::select(dep, iso_a2, gas_import_dependency_pct), by = "iso_a2")

# frame wide enough to show the study countries in full
frame <- st_bbox(st_transform(
  st_as_sfc(st_bbox(c(xmin = 6, ymin = 47, xmax = 31, ymax = 66), crs = 4326)), laea))

ggplot() +
  geom_sf(data = dep_map, aes(fill = gas_import_dependency_pct),
          colour = "white", linewidth = 0.3) +
  scale_fill_gradient(low = "#eef2f5", high = "#8f1d17", na.value = "#e9edf1",
                      name = "Import\ndependency (%)") +
  coord_sf(xlim = c(frame[["xmin"]], frame[["xmax"]]),
           ylim = c(frame[["ymin"]], frame[["ymax"]]), expand = FALSE) +
  labs(title = "Gas import dependency by country, 2023",
       subtitle = "Share of gas demand met by net imports (Eurostat nrg_ind_id)",
       caption = "Uncoloured countries are outside the study set. Values above 100% reflect stock draws or re-exports.") +
  theme_map()
```

## Import capacity by country

Not every country has its own terminal. This rolls the operating terminals up by
country, next to each country's gas import dependency.

```{r country-view}
#| echo: false
dep_tbl <- read_csv("data/processed/import_dependence.csv", show_col_types = FALSE)

by_country <- terminals |>
  st_drop_geometry() |>
  filter(status == "operating") |>
  group_by(country) |>
  summarise(terminals = dplyr::n(),
            capacity  = round(sum(capacity_bcm), 1), .groups = "drop")

dep_tbl |>
  left_join(by_country, by = "country") |>
  mutate(terminals = coalesce(terminals, 0L),
         capacity  = coalesce(capacity, 0)) |>
  arrange(desc(gas_import_dependency_pct), desc(capacity)) |>
  transmute(Country = country,
            `Operating terminals` = terminals,
            `Capacity (bcm)` = capacity,
            `Import dependency (%)` = gas_import_dependency_pct) |>
  reactable(sortable = TRUE, highlight = TRUE, borderless = TRUE,
            pagination = FALSE, theme = rtheme,
            columns = list(
              `Operating terminals`   = colDef(align = "center"),
              `Capacity (bcm)`        = colDef(align = "center"),
              `Import dependency (%)` = colDef(align = "center")))
```

Estonia, Latvia and Sweden have no terminal of their own, yet almost all their
gas is imported; they lean on neighbours through the Balticconnector and regional
pipelines. Lithuania, Germany and Poland each run a single terminal, so that one
asset carries the country's seaborne supply. Denmark stands apart, with low
dependency on the back of its North Sea production.

## Is the capacity enough?

Capacity only means something next to demand. This compares each country's
operating terminal capacity with the gas it actually burned in 2023, so a bar at
100% means the terminals alone could cover a normal year's consumption.

```{r coverage}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7.5
#| fig-height: 4.2
#| fig-alt: "Bar chart of terminal capacity as a share of national gas demand, with Finland and Lithuania far above 100 percent, Poland and Germany well below, and Estonia, Latvia, Sweden and Denmark at zero."
demand <- read_csv("data/processed/gas_demand.csv", show_col_types = FALSE)

cover <- demand |>
  left_join(by_country, by = "country") |>
  mutate(capacity = coalesce(capacity, 0),
         coverage = 100 * capacity / gas_demand_bcm)

ggplot(cover, aes(x = reorder(country, coverage), y = coverage)) +
  geom_col(aes(fill = coverage >= 100), width = 0.68) +
  geom_hline(yintercept = 100, linetype = "31", colour = ink, linewidth = 0.4) +
  geom_text(aes(label = ifelse(coverage == 0, "no terminal",
                               paste0(round(coverage), "%"))),
            hjust = -0.12, size = 3, colour = muted) +
  scale_fill_manual(values = c(`TRUE` = "#2c5d7c", `FALSE` = "#8f1d17"),
                    guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.22))) +
  coord_flip() +
  labs(x = NULL, y = "Terminal capacity as a share of national gas demand (%)",
       title = "Can the terminals cover what each country burns?",
       subtitle = "Dashed line is 100%; blue clears it, red falls short",
       caption = "Operating terminals only. Demand is inland consumption, 2023 (Eurostat nrg_cb_gas).") +
  theme_transit() +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor = element_blank())
```

```{r coverage-numbers}
#| echo: false
tot_cap <- sum(cover$capacity)
tot_dem <- sum(cover$gas_demand_bcm)
reg_cov <- round(100 * tot_cap / tot_dem)
```

Across the eight countries the operating terminals could land `r reg_cov`% of the
gas the region burns in a year. That regional figure hides the more interesting
split. Finland and Lithuania built capacity well beyond their own consumption, so
their terminals are not just national supply but a regional buffer that neighbours
can draw on. Poland and Germany, with far larger demand, cover only part of it by
sea and still lean on pipelines from the west and north.

Four countries show no terminal, but they are not in the same position. Estonia,
Latvia and Sweden import all their gas through neighbours, so the empty bar is a
dependency. Denmark's empty bar is not: it produces its own gas from the North Sea
and is the one country here that does not need a terminal.

Read alongside the exposure score, this sharpens the picture. The terminals sitting
closest to Russian territory are the small ones, but two of them, Klaipėda and
Inkoo, are precisely the assets carrying spare capacity for their neighbours. The
most exposed sites are also the ones the region's fallback depends on.

## Exposure ranking

Click any column heading to re-sort the table.

```{r ranking}
#| echo: false
rank_df <- terminals |>
  st_drop_geometry() |>
  transmute(
    Terminal   = name,
    Country    = country,
    Status     = status,
    Capacity   = round(capacity_bcm, 2),
    `km to Russia`  = round(dist_russia_km),
    `km to Straits` = round(dist_chokepoint_km),
    Exposure   = round(exposure, 2)
  )

exp_rng  <- range(rank_df$Exposure)
exp_ramp <- colorRamp(exp_cols)

reactable(
  rank_df,
  defaultSorted = list(Exposure = "desc"),
  sortable = TRUE, highlight = TRUE, borderless = TRUE, pagination = FALSE,
  columns = list(
    Capacity        = colDef(name = "Capacity (bcm)", align = "center"),
    `km to Russia`  = colDef(align = "center"),
    `km to Straits` = colDef(align = "center"),
    Exposure = colDef(align = "center", style = function(value) {
      x <- (value - exp_rng[1]) / (exp_rng[2] - exp_rng[1])
      list(background  = grDevices::rgb(exp_ramp(x), maxColorValue = 255),
           color       = if (x > 0.55) "#ffffff" else "#12263f",
           fontWeight  = 600)
    })
  ),
  theme = rtheme
)
```

## Change the weights yourself

The weights are a choice, so rather than defend one set, here they are to move.
Drag a slider or take one of the presets, and the ranking recomputes as you go.
The weights are rescaled to sum to one, so only their balance matters.

```{r weights-data}
#| echo: false
#| output: asis
wd <- st_drop_geometry(terminals) |>
  transmute(name,
            p = round(s_proximity, 4),
            c = round(s_capacity, 4),
            k = round(s_chokepoint, 4))
cat('<script>window.terminalScores = ',
    jsonlite::toJSON(wd, dataframe = "rows"), ';</script>', sep = "")
```

```{=html}
<div class="wx">
  <div class="wx-presets">
    <span>Presets:</span>
    <button type="button" data-w="40,30,30">Balanced</button>
    <button type="button" data-w="60,20,20">Proximity-led</button>
    <button type="button" data-w="20,60,20">Capacity-led</button>
    <button type="button" data-w="20,20,60">Transit-led</button>
  </div>
  <div class="wx-controls">
    <label>Proximity to Russia <output id="wx-p-out">40%</output>
      <input type="range" id="wx-p" min="0" max="100" step="5" value="40"></label>
    <label>Share of capacity <output id="wx-c-out">30%</output>
      <input type="range" id="wx-c" min="0" max="100" step="5" value="30"></label>
    <label>Route past the Straits <output id="wx-k-out">30%</output>
      <input type="range" id="wx-k" min="0" max="100" step="5" value="30"></label>
  </div>
  <table class="wx-table">
    <thead><tr><th>Rank</th><th>Terminal</th><th>Exposure</th></tr></thead>
    <tbody id="wx-rows"></tbody>
  </table>
</div>

<style>
.wx { border: 1px solid rgba(18,38,63,0.13); padding: 1rem 1.1rem; margin: 1.2rem 0; }
.wx-presets { font-size: 0.85rem; color: #5a6675; margin-bottom: 0.9rem; }
.wx-presets button {
  font: inherit; font-size: 0.82rem; color: #12263f; background: #fff;
  border: 1px solid rgba(18,38,63,0.3); padding: 0.2rem 0.6rem; margin-right: 0.3rem;
  cursor: pointer; border-radius: 2px;
}
.wx-presets button:hover { background: #eef2f6; }
.wx-presets button[aria-pressed="true"] { background: #12263f; color: #fff; border-color: #12263f; }
.wx-controls { display: grid; grid-template-columns: repeat(auto-fit, minmax(190px, 1fr)); gap: 0.9rem; }
.wx-controls label { display: block; font-size: 0.85rem; color: #24303f; }
.wx-controls output { font-weight: 600; color: #12263f; }
.wx-controls input[type=range] { width: 100%; margin-top: 0.3rem; accent-color: #8f1d17; }
.wx-table { width: 100%; margin-top: 1.1rem; font-size: 0.9rem; border-collapse: collapse; }
.wx-table th { text-align: left; border-bottom: 2px solid #12263f; padding: 0.3rem 0.4rem; }
.wx-table td { border-bottom: 1px solid rgba(18,38,63,0.13); padding: 0.35rem 0.4rem; }
.wx-table td:first-child { width: 2.6rem; color: #5a6675; font-variant-numeric: tabular-nums; }
.wx-bar { display: inline-block; height: 0.7rem; background: #8f1d17; vertical-align: middle;
  margin-right: 0.5rem; min-width: 2px; }
.wx-val { font-variant-numeric: tabular-nums; color: #24303f; }
</style>

<script>
(function () {
  var data = window.terminalScores;
  var rows = document.getElementById('wx-rows');
  if (!data || !rows) return;
  var ids = ['p', 'c', 'k'];
  var inputs = {}, outs = {};
  ids.forEach(function (i) {
    inputs[i] = document.getElementById('wx-' + i);
    outs[i]   = document.getElementById('wx-' + i + '-out');
  });

  function render() {
    var raw = {}, total = 0;
    ids.forEach(function (i) { raw[i] = Number(inputs[i].value); total += raw[i]; });
    if (total === 0) { total = 1; raw.p = 1; }
    ids.forEach(function (i) {
      outs[i].textContent = Math.round(100 * raw[i] / total) + '%';
    });
    var scored = data.map(function (d) {
      return { name: d.name,
               score: (raw.p * d.p + raw.c * d.c + raw.k * d.k) / total };
    }).sort(function (a, b) { return b.score - a.score; });

    var max = scored[0].score || 1;
    rows.innerHTML = scored.map(function (d, i) {
      var w = Math.max(2, Math.round(120 * d.score / max));
      return '<tr><td>' + (i + 1) + '</td><td>' + d.name + '</td>' +
             '<td><span class="wx-bar" style="width:' + w + 'px"></span>' +
             '<span class="wx-val">' + d.score.toFixed(2) + '</span></td></tr>';
    }).join('');

    var current = ids.map(function (i) { return raw[i]; }).join(',');
    document.querySelectorAll('.wx-presets button').forEach(function (b) {
      b.setAttribute('aria-pressed', b.dataset.w === current ? 'true' : 'false');
    });
  }

  ids.forEach(function (i) { inputs[i].addEventListener('input', render); });
  document.querySelectorAll('.wx-presets button').forEach(function (b) {
    b.addEventListener('click', function () {
      var w = b.dataset.w.split(',');
      ids.forEach(function (id, n) { inputs[id].value = w[n]; });
      render();
    });
  });
  render();
})();
</script>
```

Weighting towards proximity lifts the terminals nearest Russian territory, Hamina,
Klaipėda and Gdańsk. Weighting towards capacity lifts Świnoujście, the largest in
the region. That the order moves at all is the point: the ranking is a function of
the assumptions, and the assumptions are yours to set.

## Interactive map

Toggle the layers in the top-right, hover a corridor or terminal for its name,
and click a terminal for its figures.

```{r leaflet}
#| echo: false
terminals_wgs <- st_transform(terminals, 4326)
russia_wgs    <- st_transform(st_crop(russia, aoi), 4326)
transit_wgs   <- st_transform(transit, 4326)
choke_bb      <- st_bbox(st_transform(chokepoint, 4326))
pal <- colorNumeric(exp_cols, domain = terminals_wgs$exposure)
tb  <- st_bbox(terminals_wgs)

leaflet(terminals_wgs) |>
  addProviderTiles("CartoDB.Positron") |>
  # Focus the initial view on the terminals. Without this, leaflet auto-fits to
  # all layers, and the Russia polygon stretches the view east into empty land.
  fitBounds(tb[["xmin"]] - 1.5, tb[["ymin"]] - 0.8,
            tb[["xmax"]] + 1.5, tb[["ymax"]] + 0.8) |>
  addPolygons(data = russia_wgs, fillColor = "#8f1d17", fillOpacity = 0.16,
              weight = 0, group = "Russian territory", label = "Russian territory",
              highlightOptions = highlightOptions(fillOpacity = 0.32)) |>
  addRectangles(lng1 = choke_bb[["xmin"]], lat1 = choke_bb[["ymin"]],
                lng2 = choke_bb[["xmax"]], lat2 = choke_bb[["ymax"]],
                fill = FALSE, color = "#12263f", weight = 2, dashArray = "5,5",
                group = "Danish Straits", label = "Danish Straits") |>
  addPolylines(data = transit_wgs, color = ~unname(corridor_cols[state]), weight = 3,
               opacity = 0.9, label = ~paste0(name, " (", status, ")"),
               highlightOptions = highlightOptions(weight = 6, opacity = 1),
               group = "Transit corridors") |>
  addCircleMarkers(
    radius      = ~rescale(capacity_bcm, to = c(5, 16)),
    color       = "#12263f", weight = 1,
    fillColor   = ~pal(exposure), fillOpacity = 0.85,
    group       = "LNG terminals",
    label       = ~name,
    popup       = ~paste0("<b>", name, "</b><br>", country,
                          "<br>Capacity: ", capacity_bcm, " bcm/yr",
                          "<br>Status: ", status,
                          "<br>Exposure: ", round(exposure, 2))
  ) |>
  addLegend("bottomright", pal = pal, values = ~exposure, title = "Exposure",
            opacity = 0.9) |>
  addScaleBar(position = "bottomleft") |>
  addMiniMap(toggleDisplay = TRUE, minimized = FALSE, width = 110, height = 110) |>
  addLayersControl(
    overlayGroups = c("LNG terminals", "Transit corridors",
                      "Russian territory", "Danish Straits"),
    options = layersControlOptions(collapsed = FALSE))
```
