---
title: "Baltic Energy Transit Risk"
subtitle: "Exposure of post-2022 energy import infrastructure"
author: "Tobyn Smith"
format:
  pdf:
    pdf-engine: xelatex
    documentclass: article
    geometry:
      - margin=2.3cm
    fontsize: 10pt
    fig-pos: "H"
    colorlinks: true
    linkcolor: RoyalBlue
    urlcolor: RoyalBlue
    include-in-header:
      text: |
        \usepackage{float}
        \usepackage[default,scale=0.96]{sourceserifpro}
        \usepackage[scaled=0.92]{sourcesanspro}
        \usepackage{sectsty}
        \sectionfont{\color[HTML]{12263F}\sffamily\large}
        \subsectionfont{\color[HTML]{12263F}\sffamily}
        \renewcommand{\arraystretch}{1.15}
execute:
  warning: false
  message: false
  echo: false
---

```{r setup}
library(sf)
library(dplyr)
library(readr)
library(ggplot2)
library(rnaturalearth)
library(ggrepel)
library(knitr)

laea <- 3035
ink   <- "#12263f"
muted <- "#5a6675"
land  <- "#e6ebf0"
sea   <- "#f7f9fb"
exp_cols <- c("#e9eef2", "#cf8f6a", "#8f1d17")

theme_brief <- function() {
  theme_minimal(base_size = 9) +
    theme(plot.title    = element_text(face = "bold", colour = ink, size = 11),
          plot.subtitle = element_text(colour = muted, size = 8),
          plot.caption  = element_text(colour = muted, size = 6.5, hjust = 0),
          legend.title  = element_text(size = 7, colour = muted),
          legend.text   = element_text(size = 6.5))
}

world  <- ne_countries(scale = "medium", returnclass = "sf")
aoi    <- st_as_sfc(st_bbox(c(xmin = 9, ymin = 53, xmax = 30, ymax = 66),
                            crs = 4326)) |> st_transform(laea)
russia <- world |> filter(admin == "Russia") |> st_transform(laea) |> st_make_valid()
chokepoint <- st_as_sfc(st_bbox(c(xmin = 10.5, ymin = 55.0, xmax = 13.0, ymax = 56.2),
                                crs = 4326)) |> st_transform(laea)

terminals <- read_csv("data/processed/lng_terminals.csv", show_col_types = FALSE) |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  st_transform(laea) |>
  mutate(dist_russia_km = as.numeric(st_distance(geometry, st_union(russia))) / 1000,
         dist_chokepoint_km = as.numeric(st_distance(geometry, st_union(chokepoint))) / 1000)

norm <- function(x) (max(x, na.rm = TRUE) - x) /
                    (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
terminals <- terminals |>
  mutate(s_proximity  = norm(dist_russia_km),
         s_chokepoint = 1 - norm(dist_chokepoint_km),
         s_capacity   = capacity_bcm / max(capacity_bcm, na.rm = TRUE),
         exposure     = 0.4 * s_proximity + 0.3 * s_capacity + 0.3 * s_chokepoint) |>
  arrange(desc(exposure))

land_aoi <- world |> st_transform(laea) |> st_crop(aoi)

ru      <- read_csv("data/processed/russian_imports.csv", show_col_types = FALSE)
demand  <- read_csv("data/processed/gas_demand.csv", show_col_types = FALSE)
ru_21   <- sum(ru$russian_gas_bcm[ru$year == 2021])
ru_23   <- sum(ru$russian_gas_bcm[ru$year == 2023])
lng_cap <- sum(terminals$capacity_bcm[terminals$status == "operating"])
reg_cov <- round(100 * lng_cap / sum(demand$gas_demand_bcm))
```

## Executive summary

The Baltic region replaced a continental pipeline network with six coastal sites.
Russian pipeline gas into the eight littoral states fell from roughly
`r round(ru_21)` bcm in 2021 to `r round(ru_23, 1)` bcm in 2023, while the
terminals built to receive seaborne gas hold about `r round(lng_cap, 1)` bcm a
year, some `r round(100 * lng_cap / ru_21)`% of what stopped arriving.

The strategic problem is not the size of that gap, which lower demand and pipeline
gas from the west and north have largely closed. It is that the infrastructure
which replaced the pipelines is concentrated, sits nearest Russian territory
precisely where it is smallest, and is one asset deep in every country that holds
it. Policymakers should therefore treat these terminals less as a solved
substitution and more as a set of single points of failure whose loss would push
the region back onto the same subsea links that have already been cut three times
since 2022.

This brief sets out where that infrastructure sits, how exposed each terminal is on
a stated and adjustable measure, what happens when the largest terminal in each
country is removed, and what follows for policy.

## Where the terminals sit

```{r map, fig.width = 6.6, fig.height = 5.0}
ggplot() +
  geom_sf(data = land_aoi, fill = land, colour = "#d3dae1", linewidth = 0.18) +
  geom_sf(data = st_crop(russia, aoi), fill = "#dfd0cf", colour = "#c9b4b2",
          linewidth = 0.18) +
  geom_sf(data = st_sf(geometry = chokepoint), fill = NA, colour = ink,
          linewidth = 0.4, linetype = "31") +
  geom_sf(data = terminals, aes(size = capacity_bcm, colour = exposure)) +
  geom_text_repel(data = terminals, aes(label = name, geometry = geometry),
                  stat = "sf_coordinates", size = 2.2, colour = ink, seed = 4,
                  min.segment.length = 0, segment.colour = "#8e9aa6",
                  segment.size = 0.15, box.padding = 0.75, point.padding = 0.5,
                  force = 4, max.overlaps = Inf) +
  scale_size_continuous(range = c(1.6, 6.5), name = "Capacity\n(bcm/yr)") +
  scale_colour_gradientn(colours = exp_cols, name = "Exposure") +
  labs(title = "Baltic LNG import terminals",
       subtitle = "Point size is capacity, colour is exposure; dashed box is the Danish Straits",
       caption = "Sources: Natural Earth; operator reporting and Global Energy Monitor.") +
  theme_brief() +
  theme(axis.text = element_blank(), axis.title = element_blank(),
        panel.grid = element_blank(),
        panel.background = element_rect(fill = sea, colour = NA))
```

## Findings

**Capacity is concentrated in the south.** Świnoujście in Poland and Mukran in
Germany hold about three fifths of the region's operating import capacity. The
northeast leans on one or two terminals each, Inkoo for the Gulf of Finland and
Klaipėda for Lithuania. What replaced a continental pipeline network is, in
practice, six coastal sites.

**The terminals nearest Russia are the small ones.** Hamina, at the eastern end of
the Gulf of Finland, and Klaipėda and Gdańsk beside the Kaliningrad exclave sit
closest. The two largest terminals are among the farthest away. Size and proximity
do not line up, which is precisely what the exposure score is built to expose.

**Every cargo enters through one strait.** All LNG bound for these terminals
passes the Danish Straits. The Kiel Canal is the only other way in and is far too
small for LNG carriers. The terminals deepest inside the sea run the longest
exposed route to port.

**The small states overbuilt on purpose.** Set against what each country actually
burns, Finland and Lithuania hold capacity worth several times their own demand,
while Germany and Poland cover only part of theirs by sea. Across the eight
countries the terminals could land `r reg_cov`% of annual consumption.

```{r ranking}
terminals |>
  st_drop_geometry() |>
  transmute(Terminal = name, Country = country, Status = status,
            `Capacity (bcm)` = capacity_bcm,
            `km to Russia` = round(dist_russia_km),
            Exposure = round(exposure, 2)) |>
  kable(booktabs = TRUE, caption = "Terminals ranked by exposure score.")
```

## What follows for policy

1. **The northeast has no spare door.** Borrowing the logic of the EU's N-1
   standard and removing each country's largest terminal, Finland is left covering
   about a fifth of its demand and Lithuania, Poland and Germany are left with no
   seaborne capacity at all. They are not failing the real test, which counts
   pipelines and storage too, but the fallback is exactly the links that have been
   damaged since 2022.
2. **The connectors matter as much as the terminals.** Estonia, Latvia and Sweden
   import through neighbours rather than terminals of their own, over links such as
   the Balticconnector that have already been damaged.
3. **The Danish Straits are a shared dependency.** One entrance serves all eight
   countries, so surveillance and patrol arrangements there protect the whole
   region at once.
4. **Floating terminals cut both ways.** They were quick to add and can be moved,
   but they are few and chartered. Mukran lost half its capacity when a charter
   ended.

## This is not hypothetical

| Date | Asset | What happened |
|:-----|:------|:--------------|
| Sep 2022 | Nord Stream 1 and 2 | Pipelines ruptured by underwater explosions off Bornholm |
| Oct 2023 | Balticconnector | Pipeline and cable damaged by a ship's anchor; back in service 22 April 2024 |
| Dec 2024 | Estlink 2 | Power and telecoms cables cut; back in service 20 June 2025 |

## Limitations

This measures geographic exposure, not the probability of disruption. The score
uses proxies rather than a causal model, and different reasonable weights shift the
ranking. It is a single snapshot: capacities move, and headline figures often
describe planned rather than operating capacity. Demand in 2023 was itself
depressed by high prices and mild weather, so coverage in a colder year would look
less comfortable.

## Sources

All public. Terminal locations and capacities from operator reporting and the
Global Energy Monitor LNG tracker, current to 2024 and 2025. Gas import dependency
(`nrg_ind_id`), inland consumption (`nrg_cb_gas`) and imports from Russia
(`nrg_ti_gas`) from Eurostat. Country geography from Natural Earth.

Full data, methods and interactive maps: <https://tobyn-smith.github.io/transit>
