---
title: "Baltic Energy Transit Risk"
subtitle: "Exposure of post-2022 Baltic energy import infrastructure"
date: today
date-format: "D MMMM YYYY"
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
library(sf)
library(readr)
library(leaflet)
library(scales)

terminals <- read_csv("data/processed/lng_terminals.csv", show_col_types = FALSE) |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)
```

## In brief

After 2022 the Baltic region swapped Russian pipeline gas for seaborne LNG,
landed at a handful of coastal terminals. This project maps that infrastructure
and scores how exposed each terminal is, using distance to Russian territory,
share of import capacity, and the length of the supply run past the Danish
Straits. Three things stand out: capacity is concentrated in two southern
terminals, the terminals nearest Russia are the small ones, and every cargo
enters through a single strait. The infrastructure has already been hit several
times since 2022, so none of this is abstract.

## The question

Before 2022, much of the Baltic region's gas arrived by pipeline from Russia.
After the invasion of Ukraine and the sabotage of the Nord Stream pipelines, that
supply was cut or abandoned, and imports moved onto a small set of coastal
terminals for liquefied natural gas (LNG), such as Klaipėda, Świnoujście, and
Inkoo. These are the sites where tankers unload gas that has been chilled to a
liquid for shipping, and turn it back into gas for the pipeline network.

That shift concentrated a great deal of supply onto a handful of coastal sites.
This project asks how concentrated, and how exposed:

> After 2022, Baltic energy imports shifted onto a small set of coastal LNG
> terminals and ports. How concentrated is that infrastructure, and how exposed
> is it, measured by capacity, by distance to Russian territory, and by reliance
> on passage through the Danish Straits?

The terminals, at a glance. Hover for a name, click for the detail; each point is
sized by capacity.

```{r overview-map}
#| echo: false
leaflet(terminals) |>
  addProviderTiles("CartoDB.Positron") |>
  addCircleMarkers(
    radius      = ~rescale(capacity_bcm, to = c(5, 15)),
    color       = "#17150f", weight = 1,
    fillColor   = "#9e2b25", fillOpacity = 0.8,
    label       = ~name,
    popup       = ~paste0("<b>", name, "</b><br>", country,
                          "<br>", capacity_bcm, " bcm/yr<br>", status))
```

## How to read this site

- The [Data](01-data.qmd) page lists every source, all public, and says what is
  solid and what is still rough.
- The [Analysis](02-analysis.qmd) page holds the maps, the distance
  measurements, and the exposure ranking.
- In a hurry? The [Slides](slides.qmd) walk through the whole thing as a short,
  navigable deck.

You do not need any technical background to read it. The code behind each chart is
tucked away and can be ignored.

## A few terms

- **LNG (liquefied natural gas):** natural gas chilled to a liquid so it can be
  carried by ship rather than a pipeline.
- **LNG terminal:** the coastal site where those ships unload and the cargo is
  turned back into gas for the pipeline network. Some are built on land; some are
  a moored ship, known as an FSRU (a floating storage and regasification unit).
- **Capacity (bcm):** how much gas a terminal can bring in over a year, measured
  in billion cubic metres.
- **Import dependency:** the share of a country's gas that comes from abroad
  rather than from its own production.

## Findings

The numbers behind these sit on the [Analysis](02-analysis.qmd) page. The
terminal figures are compiled from operator reporting and Global Energy Monitor
(see the [Data](01-data.qmd) page), so the rankings show the broad pattern of
exposure, not precise measurement.

1. **Concentration.** Capacity is lopsided. Two southern terminals, Mukran in
   Germany and Świnoujście in Poland, hold most of the region's import capacity,
   while the northeast leans on one or two terminals each: Inkoo for the Gulf of
   Finland and Klaipėda for Lithuania.
2. **Proximity.** The terminals closest to Russian territory are Hamina, at the
   eastern end of the Gulf of Finland, and Klaipėda and Gdańsk, which sit beside
   the Kaliningrad exclave. Proximity and capacity do not always line up: the two
   largest terminals, Mukran and Świnoujście, are among the farthest from Russia,
   which is what the exposure score is built to show.
3. **Transit through the Straits.** Every LNG cargo bound for these terminals
   passes the Danish Straits. The Kiel Canal is the only other way in, and it is
   far too small for LNG carriers, so the Straits are the effective entrance. The
   terminals deepest inside the sea, in the Gulf of Finland, run the longest
   exposed route to reach port.

## The exposure score

Each terminal gets a score built from three things, each with a stated weight:

- how close it is to Russian territory (weight 0.4),
- how much of the region's import capacity it holds (0.3),
- how long its supply route past the Danish Straits runs (0.3).

Proximity gets the most weight because the question is about exposure to pressure
from one actor, and distance is the most direct measure the public data supports.
Capacity matters because a larger terminal does more damage if it goes down, and
transit matters because a terminal deep inside the Baltic runs a longer voyage to
be resupplied.

The weights are a judgement, not a finding. They sit in plain view on the
Analysis page, changing them changes the ranking, and the sensitivity table there
shows by how much. There is no hidden model. It is a starting point, not a final
score.

## Disruption to the infrastructure since 2022

This exposure is not hypothetical. Baltic energy and data links have been damaged
on several occasions since 2022, mostly by vessels dragging anchors across the
seabed.

| Date | Asset | What happened |
|------|-------|---------------|
| Sep 2022 | Nord Stream 1 and 2 | Subsea gas pipelines ruptured by underwater explosions off Bornholm. |
| Oct 2023 | Balticconnector | The Finland-Estonia gas pipeline and a telecoms cable were damaged by a ship's anchor; the pipeline was repaired in 2024. |
| Dec 2024 | Estlink 2 | The Estonia-Finland power cable and four telecoms cables were cut; a shadow-fleet tanker is the prime suspect. The cable was back in service in 2025. |

These incidents are context, not events modelled here, but they are why the
geography in this project is worth looking at closely. The
[Analysis](02-analysis.qmd) page maps the main corridors and which have been hit.

Sources: [Balticconnector (ERR)](https://news.err.ee/1609422658/china-admits-container-ship-newnew-polar-bear-damaged-undersea-gas-pipeline),
[Estlink 2 incident (Wikipedia)](https://en.wikipedia.org/wiki/2024_Estlink_2_incident).

## What follows for policy

Four points come out of the maps rather than out of theory:

1. **The north-east has no spare door.** Finland and Lithuania each lean on a
   single large terminal. An outage at Inkoo or Klaipėda, planned or otherwise,
   has no local substitute; backup arrives by pipeline from neighbours, over
   exactly the links that have been damaged since 2022.
2. **The connectors matter as much as the terminals.** Estonia, Latvia and Sweden
   import through neighbours rather than terminals of their own, so their supply
   runs on pipes and cables like the Balticconnector. The incidents above hit
   precisely that layer.
3. **The Danish Straits are a shared dependency.** Every cargo for all eight
   countries passes one entrance, so monitoring and patrol arrangements there, of
   the kind NATO stood up after the December 2024 cable cuts, protect the whole
   region at once.
4. **Floating terminals cut both ways.** FSRUs were quick to add and can be moved
   to where they are needed, but they are few, chartered, and easy to count. Much
   of the region's flexibility rests on a handful of ships.

## Limitations

This measures geographic exposure, not the probability of disruption. A terminal
close to a maritime boundary is not a prediction that anything will happen to it.
It is a statement about geography.

A few specific caveats:

- The score uses proxy measures (distance, capacity share) rather than a causal
  model. Different reasonable weights would shift the rankings.
- It is a single snapshot. Capacity, routes, and terminals change.
- The terminal capacities are compiled from operator reporting and Global Energy
  Monitor, current to 2024 and 2025. The dependence figures are Eurostat data for
  2023. See the [Data](01-data.qmd) page.
- Undersea cable and pipeline incidents in the region between 2023 and 2025 are
  context for the question, not events modelled here.

This shows where the exposure sits. It is not a risk forecast.
