---
title: "Baltic Energy Transit Risk"
subtitle: "How exposed the new Baltic gas import terminals are"
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
library(sf)
library(readr)
library(leaflet)
library(scales)

terminals <- read_csv("data/processed/lng_terminals.csv", show_col_types = FALSE) |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)
```

## What this is

This started as a student project in international affairs. I wanted a map of
the terminals the Baltic region built after Russian pipeline gas stopped, and
some way of saying how exposed each one looks if you only use public data.

In 2021 the eight countries on that sea took about 70 bcm from Russia by pipe.
By 2023 the same Eurostat series was down to almost nothing. Five working
coastal LNG terminals, plus a sixth planned at Gdańsk, can land about 23 bcm a
year between them, which is roughly a third of what stopped arriving. Demand
fell, and gas still comes in by pipe from Norway and the west, so the rest of
the gap closed that way rather than through these sites alone.

On a normal year, then, the region is not short of gas. What I kept coming
back to is how concentrated those terminals are, how close they sit to Russian
territory, and how much of the supply run depends on ships coming through the
Danish Straits. I score each terminal on those three things, with weights I
picked myself, so the ranking is a judgement rather than a discovered fact.
You can move the weights on the Analysis page and watch how far the order
shifts.

Nord Stream was blown in 2022, Balticconnector was damaged in 2023, and
Estlink 2 was cut in 2024. I am mapping the geography those events keep
landing on, not scoring the events themselves.

Terminal capacities are operator figures from 2024 and 2025, and import
dependency is Eurostat for 2023. Everything I used is on the
[Data](01-data.qmd) page.

## The question

Before 2022 most of this gas came by pipe from Russia. After the invasion of
Ukraine and the Nord Stream sabotage, that route was shut or abandoned, and
the region started landing chilled gas from ships at sites such as Klaipėda,
Świnoujście and Inkoo. An LNG terminal is the coastal site where those ships
unload and the cargo is warmed back into gas for the pipe network. Some of
them are built on land; others are a moored ship, an FSRU.

That move put a lot of eight countries' import capacity onto a few coastal
points you can actually put on a map. The question I am asking is:

> After 2022, Baltic energy imports shifted onto a small set of coastal LNG
> terminals. How concentrated is that infrastructure, and how exposed is it, if
> you measure by capacity, by distance to Russian territory, and by how far a
> cargo has to run past the Danish Straits?

The terminals, at a glance. Hover for a name, click for the figures. Point
size is capacity.

```{r overview-map}
#| echo: false
leaflet(terminals) |>
  addProviderTiles("CartoDB.Positron") |>
  addCircleMarkers(
    radius      = ~rescale(capacity_bcm, to = c(5, 15)),
    color       = "#12263f", weight = 1,
    fillColor   = "#8f1d17", fillOpacity = 0.8,
    label       = ~name,
    popup       = ~paste0("<b>", name, "</b><br>", country,
                          "<br>", capacity_bcm, " bcm/yr<br>", status))
```

## What I found

The numbers sit on the [Analysis](02-analysis.qmd) page. Capacities move, so I
would read them as a picture of the pattern rather than as a precise
measurement.

Świnoujście in Poland and Mukran in Germany hold about three fifths of the
operating import capacity, which already tells you how lopsided the map is.
The northeast runs on one or two terminals each: Inkoo for the Gulf of
Finland, Klaipėda for Lithuania. Five working sites replaced a continental
pipe network, with Gdańsk still planned.

The terminals closest to Russian territory are the small ones. Hamina sits at
the eastern end of the Gulf of Finland, Klaipėda and Gdańsk sit next to
Kaliningrad, and the two largest terminals are among the farthest from
Russia. I did not rank on size alone for that reason.

Every cargo still comes through the Danish Straits. The Kiel Canal is the
other way in, but it is too small for LNG carriers, so terminals up in the
Gulf of Finland have the longest run once they are inside the sea.

Finland and Lithuania built well beyond their own demand, while Germany and
Poland burn much more gas and only cover part of that by sea. The northeast,
in other words, built a buffer that neighbours draw on. That is the part I
care most about, because the terminals sitting closest to Russia are also the
ones carrying spare capacity for the rest of the region.

## How the score is built

Each terminal gets a number from three ingredients:

- how close it is to Russian territory (I gave this 0.4, the largest share),
- how much of the region's import capacity it holds (0.3),
- how long its supply route past the Danish Straits runs (0.3).

I weighted proximity most because the question is about exposure to one
neighbour, and distance is the most direct thing the public data will
support. Capacity is in there because a larger terminal does more damage if
it stops, and transit is in there because a site deep in the Gulf of Finland
has a longer voyage to be restocked.

Those weights are a judgement, written out on the Analysis page so you can
drag them and see how much the order moves. There is no hidden model; every
number can be explained out loud.

## Disruptions since 2022

Baltic energy and data links have been damaged several times since 2022.
Most of that damage has come from ships dragging anchors across the seabed.

| Date | Asset | What happened |
|------|-------|---------------|
| Sep 2022 | Nord Stream 1 and 2 | Subsea gas pipelines ruptured by underwater explosions off Bornholm. |
| Oct 2023 | Balticconnector | The Finland-Estonia gas pipeline and a telecoms cable were damaged by a ship's anchor. The pipeline was shut for six months and returned to commercial use on 22 April 2024. |
| Dec 2024 | Estlink 2 | The Estonia-Finland power cable and four telecoms cables were cut on Christmas Day; a shadow-fleet tanker is the prime suspect. Repairs ran into 2025 and the link returned to commercial use on 20 June 2025. |

I left those events as context; they are why the geography is worth looking
at. The [Analysis](02-analysis.qmd) page maps the corridors and which of them
have been hit.

Sources: [Balticconnector return to service (Gasgrid)](https://gasgrid.fi/en/2024/04/19/balticconnector-in-commercial-use-from-monday-22-april-onwards/),
[Estlink 2 return to service (Fingrid)](https://www.fingrid.fi/en/news/news/2025/estlink-2-electricity-transmission-link-returns-to-commercial-use/),
[Estlink 2 incident (Wikipedia)](https://en.wikipedia.org/wiki/2024_Estlink_2_incident).

## What I would take from this

If you borrow the EU's N-1 idea and knock out each country's largest terminal,
Finland is left with Hamina, covering about a fifth of demand, and Lithuania,
Poland and Germany are left with no seaborne capacity at all. The real N-1
test also counts pipes and storage, so they are not failing that test; the
fallback is the same links that have already been damaged.

Estonia, Latvia and Sweden have no terminal of their own. Their gas comes
through neighbours, over pipes and cables such as the Balticconnector, and
that layer has already been hit.

The Danish Straits are one entrance for all eight countries, so anything that
watches that stretch (the kind of patrol NATO stood up after the December 2024
cable cuts) covers the whole region at once.

FSRUs were quick to bring in and can be moved, but they are also few and
chartered. Mukran lost about half its capacity when a second ship came off
charter, which is why I count it at about 6 bcm rather than the 13.5 bcm still
quoted in a lot of headlines.

## Limitations

This is geographic exposure, not a probability that anything will happen. A
terminal close to a maritime boundary is a fact about a map.

A few caveats I would want a reader to have:

- The score uses distance and capacity share as proxies, so change the weights
  and the ranking moves.
- It is one snapshot: capacities, routes and terminals change.
- Capacities are compiled from operator reporting and Global Energy Monitor,
  current to 2024 and 2025. Dependence figures are Eurostat for 2023. See
  [Data](01-data.qmd).
- The 2023 to 2025 cable and pipeline incidents are context, not events in
  the model.

I would treat this as a way to see where the exposure sits, not as something
to cite as a forecast.

## Terms

- **LNG (liquefied natural gas):** natural gas chilled to a liquid so it can be
  carried by ship rather than a pipeline.
- **LNG terminal:** the coastal site where those ships unload and the cargo is
  turned back into gas for the pipeline network. Some are built on land; some are
  a moored ship, known as an FSRU (a floating storage and regasification unit).
- **Capacity (bcm):** how much gas a terminal can bring in over a year, measured
  in billion cubic metres.
- **Import dependency:** the share of a country's gas that comes from abroad
  rather than from its own production.

## The rest of the site

- [Data](01-data.qmd) lists every source, all public, and says what is solid
  and what is still rough.
- [Analysis](02-analysis.qmd) has the maps, the distances, and the ranking you
  can re-weight.
- [Slides](slides.qmd) is the same argument as a short deck.
- The [printable brief](brief.pdf) is two pages.

You do not need any technical background; the code behind each chart is
folded away.
