---
title: "Baltic Energy Transit Risk"
subtitle: "How exposed the new Baltic gas import terminals are"
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
library(sf)
library(readr)
library(leaflet)
library(scales)

terminals <- read_csv("data/processed/lng_terminals.csv", show_col_types = FALSE) |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)
```

## What this is

This started as a student project in international affairs. I wanted a map of
the terminals the Baltic region stood up after Russian pipeline gas stopped, and
a way to say how exposed each one looks from public data alone.

In 2021 the eight countries on that sea took about 70 bcm from Russia by pipe.
By 2023 the same Eurostat series was down to almost nothing. Five working
coastal LNG terminals, plus a sixth planned at Gdańsk, can land about 23 bcm a
year between them, which is roughly a third of what stopped arriving. The rest
of the gap closed because demand fell and because gas still comes in by pipe
from Norway and the west.

So the region is not short of gas on a normal year. What I wanted to see is how
concentrated those terminals are, how close they sit to Russian territory, and
how much of the supply run depends on ships coming through the Danish Straits.
I score each terminal on those three things. The weights are a choice I made,
not a result, and you can move them on the Analysis page.

Nord Stream was blown in 2022, Balticconnector was damaged in 2023, and Estlink 2
was cut in 2024. I am mapping the geography those events keep landing on, not
scoring the events themselves.

Terminal capacities are operator figures from 2024 and 2025. Import dependency is
Eurostat for 2023. Everything I used is on the [Data](01-data.qmd) page.

## The question

Before 2022 most of this gas came by pipe from Russia. After the invasion of
Ukraine and the Nord Stream sabotage, that route was shut or abandoned, and the
region started landing chilled gas from ships at sites such as Klaipėda,
Świnoujście and Inkoo. An LNG terminal is the place those ships unload, and
where the cargo is warmed back into gas for the pipe network. Some are built on
land. Some are a moored ship, an FSRU.

That move put a lot of eight countries' import capacity onto a few coastal
points you can actually put on a map. The question I am asking is:

> After 2022, Baltic energy imports shifted onto a small set of coastal LNG
> terminals. How concentrated is that infrastructure, and how exposed is it, if
> you measure by capacity, by distance to Russian territory, and by how far a
> cargo has to run past the Danish Straits?

The terminals, at a glance. Hover for a name, click for the figures. Point size
is capacity.

```{r overview-map}
#| echo: false
leaflet(terminals) |>
  addProviderTiles("CartoDB.Positron") |>
  addCircleMarkers(
    radius      = ~rescale(capacity_bcm, to = c(5, 15)),
    color       = "#12263f", weight = 1,
    fillColor   = "#8f1d17", fillOpacity = 0.8,
    label       = ~name,
    popup       = ~paste0("<b>", name, "</b><br>", country,
                          "<br>", capacity_bcm, " bcm/yr<br>", status))
```

## What I found

The numbers sit on the [Analysis](02-analysis.qmd) page. Read them as a picture
of the pattern, not as a precise measurement, because capacities move.

Capacity is lopsided. Świnoujście in Poland and Mukran in Germany hold about
three fifths of the operating import capacity. The northeast runs on one or two
terminals each: Inkoo for the Gulf of Finland, Klaipėda for Lithuania. Five
working sites replaced a continental pipe network, with Gdańsk still planned.

The terminals closest to Russian territory are the small ones. Hamina sits at
the eastern end of the Gulf of Finland; Klaipėda and Gdańsk sit next to
Kaliningrad. The two largest terminals are among the farthest from Russia. That
mismatch is why I did not rank on size alone.

Every cargo still comes through the Danish Straits. The Kiel Canal is the other
door, and it is too small for LNG carriers. Terminals up in the Gulf of Finland
have the longest run once they are inside the sea.

Finland and Lithuania built well beyond their own demand. Germany and Poland,
with much larger consumption, cover only part of theirs by sea. The northeast
built a buffer that neighbours draw on. That is the finding I would actually
take into a conversation, because the terminals sitting closest to Russia are
also the ones carrying spare capacity for the rest of the region.

## How the score is built

Each terminal gets a number from three ingredients:

- how close it is to Russian territory (I gave this 0.4, the largest share),
- how much of the region's import capacity it holds (0.3),
- how long its supply route past the Danish Straits runs (0.3).

I weighted proximity most because the question is about exposure to one
neighbour, and distance is the most direct thing the public data will support.
Capacity is in there because a larger terminal does more damage if it stops.
Transit is in there because a site deep in the Gulf of Finland has a longer
voyage to be restocked.

Those weights are a judgement. They are written out on the Analysis page, you
can drag them, and a table there shows how much the order moves. There is no
hidden model. Every number can be explained out loud.

## Disruptions since 2022

Baltic energy and data links have been damaged several times since 2022, mostly
by ships dragging anchors across the seabed.

| Date | Asset | What happened |
|------|-------|---------------|
| Sep 2022 | Nord Stream 1 and 2 | Subsea gas pipelines ruptured by underwater explosions off Bornholm. |
| Oct 2023 | Balticconnector | The Finland-Estonia gas pipeline and a telecoms cable were damaged by a ship's anchor. The pipeline was shut for six months and returned to commercial use on 22 April 2024. |
| Dec 2024 | Estlink 2 | The Estonia-Finland power cable and four telecoms cables were cut on Christmas Day; a shadow-fleet tanker is the prime suspect. Repairs ran into 2025 and the link returned to commercial use on 20 June 2025. |

I left those events as context, which is why the geography is worth looking at.
The [Analysis](02-analysis.qmd) page maps the corridors and which have been hit.

Sources: [Balticconnector return to service (Gasgrid)](https://gasgrid.fi/en/2024/04/19/balticconnector-in-commercial-use-from-monday-22-april-onwards/),
[Estlink 2 return to service (Fingrid)](https://www.fingrid.fi/en/news/news/2025/estlink-2-electricity-transmission-link-returns-to-commercial-use/),
[Estlink 2 incident (Wikipedia)](https://en.wikipedia.org/wiki/2024_Estlink_2_incident).

## What I would take from this

If you borrow the EU's N-1 idea and knock out each country's largest terminal,
Finland is left with Hamina, covering about a fifth of demand. Lithuania, Poland
and Germany are left with no seaborne capacity at all. They are not failing the
real N-1 test, which also counts pipes and storage. The fallback is the same
links that have already been damaged.

Estonia, Latvia and Sweden have no terminal of their own. Their gas comes
through neighbours, over pipes and cables such as the Balticconnector. That
layer has already been hit.

The Danish Straits are one entrance for all eight countries, so anything that
watches that stretch (the kind of patrol NATO stood up after the December 2024
cable cuts) covers the whole region at once.

FSRUs were quick to bring in and can be moved. They are also few and chartered.
Mukran lost about half its capacity when a second ship came off charter, which
is why I count it at about 6 bcm rather than the 13.5 bcm still quoted in a lot
of headlines.

## Limitations

This is geographic exposure, not a probability that anything will happen. A
terminal close to a maritime boundary is a fact about a map.

A few caveats I would want a reader to have:

- The score uses distance and capacity share as proxies. Change the weights and
  the ranking moves.
- It is one snapshot. Capacities, routes and terminals change.
- Capacities are compiled from operator reporting and Global Energy Monitor,
  current to 2024 and 2025. Dependence figures are Eurostat for 2023. See
  [Data](01-data.qmd).
- The 2023 to 2025 cable and pipeline incidents are context, not events in the
  model.

I would treat this as a way to see where the exposure sits, not as a forecast.

## Terms

- **LNG (liquefied natural gas):** natural gas chilled to a liquid so it can be
  carried by ship rather than a pipeline.
- **LNG terminal:** the coastal site where those ships unload and the cargo is
  turned back into gas for the pipeline network. Some are built on land; some are
  a moored ship, known as an FSRU (a floating storage and regasification unit).
- **Capacity (bcm):** how much gas a terminal can bring in over a year, measured
  in billion cubic metres.
- **Import dependency:** the share of a country's gas that comes from abroad
  rather than from its own production.

## The rest of the site

- [Data](01-data.qmd) lists every source, all public, and says what is solid and
  what is still rough.
- [Analysis](02-analysis.qmd) has the maps, the distances, and the ranking you
  can re-weight.
- [Slides](slides.qmd) is the same argument as a short deck.
- The [printable brief](brief.pdf) is two pages.

You do not need any technical background. The code behind each chart is folded
away.
