---
title: "Baltic Energy Transit Risk"
subtitle: "Exposure of post-2022 energy import infrastructure"
author: "Tobyn Smith"
format:
  revealjs:
    theme: [default, slides.scss]
    slide-number: c/t
    transition: fade
    progress: true
    controls: true
    hash: true
    center: false
    width: 1100
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
library(sf)
library(readr)
library(leaflet)
library(scales)

terminals <- read_csv("data/processed/lng_terminals.csv", show_col_types = FALSE) |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)
```

## The question

Before 2022, most of the Baltic region's gas came by pipeline from Russia.

After the invasion of Ukraine and the sabotage of the Nord Stream pipelines, that
supply was cut. What replaced a continental pipeline network is, in practice, six
coastal terminals for liquefied natural gas (LNG).

**How concentrated is that infrastructure, and how exposed is it?**

## The size of the hole

Russian pipeline gas into these eight countries:

- **2021:** about 70 bcm
- **2023:** about 0.3 bcm, essentially nothing

Operating LNG capacity built in the region: **around 23 bcm**, roughly a third of
what stopped arriving. The rest closed through lower demand and pipeline gas from
the west and north.

## The terminals

```{r}
#| echo: false
leaflet(terminals, height = 480) |>
  addProviderTiles("CartoDB.Positron") |>
  addCircleMarkers(
    radius      = ~rescale(capacity_bcm, to = c(6, 18)),
    color       = "#12263f", weight = 1,
    fillColor   = "#8f1d17", fillOpacity = 0.8,
    label       = ~name,
    popup       = ~paste0("<b>", name, "</b><br>", country,
                          "<br>", capacity_bcm, " bcm/yr<br>", status))
```

::: {style="font-size:0.7em;color:#5a6675"}
Hover for a name, click for the detail. Each point is sized by capacity.
:::

## What the project measures

An **exposure score** for each terminal, from three plain ingredients:

- how close it sits to Russian territory,
- how much of the region's import capacity it holds,
- how long its supply route past the Danish Straits runs.

The weights are a stated judgement, not a result. Anyone can change them.

## Finding 1: concentration

Import capacity is lopsided.

- Two southern terminals, **Świnoujście** (Poland) and **Mukran** (Germany), hold
  most of the region's capacity.
- The northeast leans on one or two terminals each: Inkoo for the Gulf of Finland,
  Klaipėda for Lithuania.

## Finding 2: proximity

The terminals closest to Russia are the **small** ones:

- **Hamina**, at the eastern end of the Gulf of Finland,
- **Klaipėda** and **Gdańsk**, beside the Kaliningrad exclave.

The two largest terminals sit farthest from Russia. Size and proximity do not line
up, which is what the score is built to show.

## Finding 3: one way in

Every LNG cargo passes the **Danish Straits**.

The Kiel Canal is the only other entrance, and it is far too small for LNG
carriers, so the Straits are the effective door.

Terminals deep in the Gulf of Finland run the longest exposed route to reach port.

## This is not hypothetical

| Date | Asset | What happened |
|------|-------|---------------|
| Sep 2022 | Nord Stream 1 & 2 | Pipelines sabotaged off Bornholm |
| Oct 2023 | Balticconnector | Pipeline and cable damaged by an anchor; repaired 2024 |
| Dec 2024 | Estlink 2 | Power and telecoms cables cut; back in service 2025 |

Baltic energy and data links have been hit repeatedly, mostly by ships dragging
anchors across the seabed.

## Who leans on whom

Not every country has its own terminal.

- **Estonia, Latvia, Sweden:** no terminal; they depend on neighbours through the
  Balticconnector and regional pipelines.
- **Lithuania, Germany, Poland:** a single terminal each, one point of failure.
- **Denmark:** low dependency, on the back of its own North Sea gas.

## Is the capacity enough?

Capacity only means something next to demand.

- **Finland and Lithuania** hold several times their own annual consumption. They
  built a regional buffer, not just national supply.
- **Germany and Poland** cover only part of much larger demand by sea, and still
  lean on pipelines from the west and north.
- **Estonia, Latvia, Sweden** have no terminal at all.

The most exposed terminals are also the ones carrying the region's spare capacity.

## Lose one terminal

Applying the logic of the EU's N-1 standard to seaborne capacity alone:

- **Finland** falls from over three times its demand to about a fifth, on Hamina
  alone.
- **Lithuania, Poland and Germany** fall to zero seaborne capacity.

They are not failing the real N-1 test, which counts pipelines and storage too.
But the fallback is precisely the pipelines and cables that have been cut.

## What follows for policy

- **No spare door in the north-east:** Finland and Lithuania each lean on one
  terminal; the fallback is the very pipelines that have been hit.
- **The connectors are the soft layer:** Estonia, Latvia and Sweden import
  through neighbours, not terminals of their own.
- **The Straits are shared:** one entrance for all eight countries, so patrols
  there protect everyone at once.
- **FSRUs cut both ways:** quick to add and movable, but few and chartered.

## Limitations

- This measures **geographic exposure**, not the probability of disruption.
- The score uses proxies (distance, capacity share), not a causal model.
- It is a single snapshot; capacity, routes and terminals change.

It shows where the exposure sits. It is not a risk forecast.

## Explore the full project

The maps, the data, and the exposure ranking, all built from public sources:

[**tobyn-smith.github.io/transit**](https://tobyn-smith.github.io/transit/)

Code and open issues on [GitHub](https://github.com/tobyn-smith/transit).
